---
title: "Example analysis of Rayan's project"
output: html_notebook
---


Read in the data

```{r}
# Load necessary libraries

library(dplyr)
library(data.table)
library(tidyr)
library(stringr)
library(DescTools)
library(ggplot2)
library(stats)
library(readr)
library(ontologyIndex)
library(readxl)
library(openxlsx)
library(boot)
library(LorenzRegression)


# a <- fread("~/Downloads/gwas_catalog_v1.0.2.1-studies_r2024-03-11.tsv")

a <- fread("/Users/rayanaloliet/Desktop/GWAS Catalog/gwas_catalog_v1.0.2.1-studies_r2024-06-07.csv")
 


```



Have a look at the data

```{r}
hist(a$`ASSOCIATION COUNT`, breaks=100)
```

Sometimes there is more than one EFO term (comma separated). Split them into separate lines

```{r}

head(a, 10)

a <- a %>%
    tidyr::separate_rows(MAPPED_TRAIT_URI, sep = ", ")

head(a, 10)

```

Get the number of EFO terms per publication

```{r}
pubmed_count <- a %>% group_by(PUBMEDID) %>% 
    summarise(
        n_efo = length(unique(MAPPED_TRAIT_URI))
    ) %>% arrange(desc(n_efo))

pubmed_count

 
```

Add the EFO count per PUBMEDID to the data frame

```{r}
a <- left_join(a, pubmed_count, by="PUBMEDID")
str(a)
```

Step 1. Develop the attention scores

- n = number of studies with that EFO term
- weighted_n = weighted by number of EFO terms in the study in total
- nhits = number of GWAS hits reported for that EFO term
- weighted_nhits = nhits weighted by number of EFO terms in the study in total

```{r}
attention <- a %>% group_by(
    MAPPED_TRAIT_URI
) %>% 
    summarise(
        n = n(), 
        weighted_n = sum(1/n_efo),
        nhits = sum(`ASSOCIATION COUNT`),
        weighted_nhits = sum(`ASSOCIATION COUNT` / n_efo)
    )

str(attention)

```

Removing extra whitespace, cleaning the Impact factor values and ensure the values are numeric 

```{r}
# Remove extra whitespace
a$`Impact factor` <- trimws(a$`Impact factor`)

# Replace non-standard characters
a$`Impact factor` <- gsub("·", ".", a$`Impact factor`)

# Convert to numeric, forcing non-numeric values to NA (for those with no impact factors)
a$`Impact factor` <- as.numeric(a$`Impact factor`)

# Replace NA values with 0
a$`Impact factor`[is.na(a$`Impact factor`)] <- 0

# Verify the changes
str(a)
unique(a$`Impact factor`)
summary(a$`Impact factor`)


```

 Including the impact factor values after preparing and cleaning the attention scores. Additionally, we have added the PUBMEDID to ensure that the studies are independent when we sum up the "n." We also added DISEASE/TRAIT for matching via str_detect function if we did not find a match via EFO terms. 
 
```{r}


# Including the impact factor and (`DISEASE/TRAIT`) column for matching and PUBMEDID to ensure the the studies are independent during the mapping process 

attention <- a %>%
  group_by(MAPPED_TRAIT_URI) %>%
  summarise(
    n = n(), 
    weighted_n = sum(1 / n_efo),
    nhits = sum(`ASSOCIATION COUNT`),
    weighted_nhits = sum(`ASSOCIATION COUNT` / n_efo),
    weighted_attention_score_impact_factor = sum((1 / n_efo) * `Impact factor`, na.rm = TRUE),
     DISEASE_TRAIT = first(`DISEASE/TRAIT`),
    PUBMEDID = paste(unique(PUBMEDID), collapse = ", ")
  )

# Expand PUBMEDID into separate rows
attention <- attention %>%
  separate_rows(PUBMEDID, sep = ", ") 


str(attention)

```
 
Step 2: Merging the GWAS attention scores with the GBD disease burden results. The GBD terms were split into first and second parts because the terms in the second part have descendants. Therefore, those in the second part need to be handled differently, where we have to pull out the descendants from the tree via ontologyIndex library and then do the matching.


```{r}

#First_part_GBD<- fread("/Users/rayanaloliet/Desktop/GWAS Catalog/GBD.csv")
First_part_GBD<- read_xlsx("/Users/rayanaloliet/Desktop/GWAS Catalog/First_part_GBD.xlsx")

str(First_part_GBD)

  # unique matched rows  by GBD term
First_part_GBD<- First_part_GBD%>%
  distinct(`GBD term`, .keep_all = TRUE)

 
```


Specify columns that contain EFO terms

```{r}

efo_columns <- c("EFO 1", "EFO 2", "EFO 3", "EFO 4", "EFO 5", "EFO 6", 
                 "EFO 7", "EFO 8", "EFO 9", "EFO 10", "EFO 11", "EFO 12",
                 "EFO 13", "EFO 14", "EFO 15", "EFO 16", "EFO 17", "EFO 18",
                 "EFO 19", "EFO 20", "EFO 21", "EFO 22", "EFO 23", "EFO 24",
                 "EFO 25", "EFO 26", "EFO 27", "EFO 28", "EFO 29", "EFO 30")
```


Reshape the GBD data to gather all relevant columns into key-value pairs and drop rows with NA EFO terms

```{r}

 gbd_long <- First_part_GBD%>%
pivot_longer(cols = all_of(efo_columns), names_to = "EFO_number", values_to =  "MAPPED_TRAIT_URI") %>% drop_na(MAPPED_TRAIT_URI) 

gbd_long
```

Separate rows based on a separator (assuming comma-separated EFO terms) and Trim whitespace

```{r}
gbd_long <- gbd_long %>%
  separate_rows(MAPPED_TRAIT_URI, sep = ",") %>%
  mutate(MAPPED_TRAIT_URI = trimws(MAPPED_TRAIT_URI))  

gbd_long
```

Ensure consistent formatting, cleaning and standardize MAPPED_TRAIT_URI in gbd_long (GBD file)

*There are many functions here because each time I attempted to find a matching, I was unable to do so due to the differences in the two datasets. For example, in GWAS, the identifiers have underscores, while in GBD they have colons, and so on... Therefore, the functions below are intended to unify these differences. 

```{r}

# Ensure consistent formatting
gbd_long$MAPPED_TRAIT_URI <- tolower(trimws(gbd_long$MAPPED_TRAIT_URI))

# Ensure lowercase
gbd_long$MAPPED_TRAIT_URI <- tolower(gbd_long$MAPPED_TRAIT_URI) 

# Trim leading/trailing whitespace
gbd_long$MAPPED_TRAIT_URI <- trimws(gbd_long$MAPPED_TRAIT_URI) 

# Remove all spaces
gbd_long$MAPPED_TRAIT_URI <- gsub("\\s+", "", gbd_long$MAPPED_TRAIT_URI) 

# Remove all internal spaces explicitly
gbd_long$MAPPED_TRAIT_URI <- gsub(" ", "", gbd_long$MAPPED_TRAIT_URI) 

# Convert to lowercase and remove spaces 
gbd_long$MAPPED_TRAIT_URI <- tolower(gsub("\\s+", "", gbd_long$MAPPED_TRAIT_URI)) 

# Remove colons and spaces
gbd_long$MAPPED_TRAIT_URI <- gsub("[:\\s]", "", gbd_long$MAPPED_TRAIT_URI)  

# Remove quotation marks
gbd_long$MAPPED_TRAIT_URI <- gsub("\"", "", gbd_long$MAPPED_TRAIT_URI) 

# Replace _ with : 
gbd_long$MAPPED_TRAIT_URI <- gsub("_", ":", gbd_long$MAPPED_TRAIT_URI) 

# Remove all extra spaces
gbd_long$MAPPED_TRAIT_URI <- str_squish(gbd_long$MAPPED_TRAIT_URI) 

(head(gbd_long$MAPPED_TRAIT_URI, 20))

```

Ensure consistent formatting, cleaning and standardize MAPPED_TRAIT_URI in attention (GWAS file) and remove rows with NA in MAPPED_TRAIT_URI (EFO terms)

```{r}
# Function to clean and standardize MAPPED_TRAIT_URI
clean_mapped_trait_uri <- function(uri) {
  # Convert to lowercase
  uri <- tolower(uri)
  # Trim leading/trailing whitespace
  uri <- trimws(uri)
  # Remove all internal spaces explicitly
  uri <- gsub("\\s+", "", uri)
  # Remove quotation marks
  uri <- gsub("\"", "", uri)
  # Replace _ with :
  uri <- gsub("_", ":", uri)
  # Remove colons and spaces
  uri <- gsub("[:\\s]", "", uri)
  # Remove prefixes and ensure consistency
  uri <- sub(".*[:/]", "", uri)
  return(uri)
}

# Apply the function to MAPPED_TRAIT_URI column
attention$MAPPED_TRAIT_URI <- clean_mapped_trait_uri(attention$MAPPED_TRAIT_URI)


# Convert "na" strings to actual NA values
attention$MAPPED_TRAIT_URI[attention$MAPPED_TRAIT_URI == "na"] <- NA

# Remove rows with NA in MAPPED_TRAIT_URI
attention <- attention %>%
  filter(!is.na(MAPPED_TRAIT_URI))

(head(attention$MAPPED_TRAIT_URI, 20))

```

Finding a match between GWAS and GBD (First_part_GBD) via identifiers (EFO terms)- for those with no descendants. Then grouping those terms by GBD term and PUBMEDID (as we need to ensure the studies are independent when we sum the n ).  

*many-to-many relationship between `x` and `y` are expected due to grouping via GBD term and PUBMEDID.

```{r}

# Merge the data frames
merged_data <- merge(attention, gbd_long, by = "MAPPED_TRAIT_URI", all.x = TRUE)

# Identify matched rows
matched_rows <- inner_join(attention, gbd_long, by = "MAPPED_TRAIT_URI")

 
 # Group by GBD term and PUBMEDID and summarize within these groups 
 matched_rows_unique <- matched_rows %>%
    group_by(`GBD term`, PUBMEDID) %>%
    summarise(
      total_attention_score_per_pubmed = first(n, na.rm = TRUE),
      weighted_n = first(weighted_n),
      nhits = first(nhits),
      weighted_nhits = first(weighted_nhits),
      weighted_attention_score_impact_factor = first(weighted_attention_score_impact_factor),
      unique_pubmed_ids = n_distinct(PUBMEDID),
      MAPPED_TRAIT_URI = first(MAPPED_TRAIT_URI),
      DISEASE_TRAIT = first(DISEASE_TRAIT),
      .groups = "drop"
    )
 
      # Now summarize these attention scores across the GBD term and sum the total_attention_score_per_pubmed 
  unique_matched_gbd_terms_First_part_GBD <- matched_rows_unique %>%
    group_by(`GBD term`) %>%
    summarise(
      total_attention_score = sum(total_attention_score_per_pubmed, na.rm = TRUE),
      weighted_n = first(weighted_n),
      nhits = first(nhits),
      weighted_nhits = first(weighted_nhits),
      weighted_attention_score_impact_factor = first(weighted_attention_score_impact_factor),
      unique_pubmed_ids = n_distinct(PUBMEDID),
      MAPPED_TRAIT_URI = first(MAPPED_TRAIT_URI),
      DISEASE_TRAIT = first(DISEASE_TRAIT),
      .groups = "drop"
    )
 
   
# Identify unmatched rows from attention
unmatched_attention <- anti_join(attention, gbd_long, by = "MAPPED_TRAIT_URI")

# Identify unmatched rows from gbd_long
unmatched_gbd <- anti_join(gbd_long, attention, by = "MAPPED_TRAIT_URI")

# Ensure unique unmatched_gbd rows by GBD term
unmatched_gbd_unique <- unmatched_gbd %>%
  distinct(`GBD term`, .keep_all = TRUE)

# Extract unique GBD terms for matched and unmatched entries
matched_gbd_terms <- matched_rows_unique$`GBD term` %>% unique()
unmatched_gbd_terms <- unmatched_gbd_unique$`GBD term` %>% unique()

# Remove any GBD terms from unmatched_gbd that are also in matched_rows
unique_unmatched_gbd_terms <- setdiff(unmatched_gbd_terms, matched_gbd_terms)

# Ensure unique unmatched_gbd rows by GBD term
unique_unmatched_gbd_terms_First_part_GBD <- unmatched_gbd_unique %>%
  filter(`GBD term` %in% unique_unmatched_gbd_terms)

# Save the data frames to Excel files
#write.xlsx(unique_matched_gbd_terms_First_part_GBD, "unique_matched_gbd_terms_First_part_GBD.xlsx")
#write.xlsx(unique_unmatched_gbd_terms_First_part_GBD, "unique_unmatched_gbd_terms_First_part_GBD.xlsx")

```


Verification of the summing up of n in total_attention_score_per_pubmed "as long as they are independent publications" (all PUBMEDID are unique)

```{r}

# Select a specific GBD term for manual verification
selected_gbd_term <- "Acne vulgaris"  # Example term

# Filter the intermediate summarization for the selected GBD term
filtered_intermediate <- matched_rows_unique %>%
  filter(`GBD term` == selected_gbd_term)

# Print the filtered data for manual verification
print(filtered_intermediate)

# Sum the `n` values manually for verification
manual_sum_n <- sum(filtered_intermediate$total_attention_score_per_pubmed)
print(paste("Manual sum of n for", selected_gbd_term, ":", manual_sum_n))

```


Finding a match between GWAS and GBD (Second_part_GBD) via identifiers- for those with  descendants. We need to change the format of the identifiers again to get the descendants using the ontology from the tree then reverse it again to match it with GWAS (to ensure the consistency in the matching process). Also, as we did not part one, we will group them by GBD term and PUBMEDID. 


```{r}


# Define utility functions
format_identifier <- function(identifier) {
  formatted <- gsub(" ", "", identifier) # Remove any spaces
  formatted <- toupper(formatted) # Convert to uppercase
  formatted <- gsub("([A-Z]+)([0-9]+)", "\\1:\\2", formatted) # Add colon
  return(formatted)
}

revert_identifier <- function(identifier) {
  reverted <- gsub(":", "", identifier) # Remove colons
  return(tolower(reverted)) # Convert to lowercase
}

clean_mapped_trait_uri <- function(uri) {
  uri <- tolower(uri) # Convert to lowercase
  uri <- trimws(uri) # Trim leading/trailing whitespace
  uri <- gsub("\\s+", "", uri) # Remove all internal spaces explicitly
  uri <- gsub("\"", "", uri) # Remove quotation marks
  uri <- gsub("_", ":", uri) # Replace _ with :
  uri <- gsub("[:\\s]", "", uri) # Remove colons and spaces
  uri <- sub(".*[:/]", "", uri) # Remove prefixes and ensure consistency
  return(uri)
}

# Read data files
ontology <- get_OBO("/Users/rayanaloliet/Desktop/GWAS Catalog/efo-obo.txt")
Second_part_GBD <- read_excel("/Users/rayanaloliet/Desktop/GWAS Catalog/Second_part_GBD.xlsx")


# Extract and combine identifiers from Second_part_GBD
identifiers_list <- list(
  Second_part_GBD$`MAPPED_TRAIT_URI...2`,
  Second_part_GBD$`MAPPED_TRAIT_URI...3`,
  Second_part_GBD$`MAPPED_TRAIT_URI...4`,
  Second_part_GBD$`MAPPED_TRAIT_URI...5`,
  Second_part_GBD$`MAPPED_TRAIT_URI...6`,
  Second_part_GBD$`MAPPED_TRAIT_URI...7`,
  Second_part_GBD$`MAPPED_TRAIT_URI...8`,
  Second_part_GBD$`MAPPED_TRAIT_URI...9`,
  Second_part_GBD$`MAPPED_TRAIT_URI...10`,
  Second_part_GBD$`MAPPED_TRAIT_URI...11`,
  Second_part_GBD$`MAPPED_TRAIT_URI...12`,
  Second_part_GBD$`MAPPED_TRAIT_URI...13`
)
identifiers <- do.call(c, lapply(identifiers_list, na.omit))

# Format the identifiers
formatted_identifiers <- sapply(identifiers, format_identifier)

# Extract descendants for each identifier
descendants_list <- list()
for (i in 1:length(formatted_identifiers)) {
  id <- formatted_identifiers[i]
  if (id %in% ontology$id) {
    descendants <- get_descendants(ontology, id)
    descendants_list[[id]] <- descendants
  } else {
    descendants_list[[id]] <- NA
  }
}

# Determine the maximum number of descendants
max_descendants <- max(sapply(descendants_list, function(x) if (is.null(x) || all(is.na(x))) 0 else length(x)), na.rm = TRUE)

# Create a data frame for descendants
descendants_df <- data.frame(MAPPED_TRAIT_URI = character(), stringsAsFactors = FALSE)
for (i in 1:max_descendants) {
  descendants_df[paste0("Descendant_", i)] <- character()
}

# Populate the data frame with identifiers and their descendants
for (i in seq_along(formatted_identifiers)) {
  id <- formatted_identifiers[i]
  descendants <- descendants_list[[id]]
  if (is.null(descendants) || all(is.na(descendants))) {
    row <- c(MAPPED_TRAIT_URI = id, rep("", max_descendants))
  } else {
    row <- c(MAPPED_TRAIT_URI = id, descendants, rep("", max_descendants - length(descendants)))
  }
  
  row_df <- as.data.frame(t(row), stringsAsFactors = FALSE)
  colnames(row_df) <- colnames(descendants_df)
  descendants_df <- rbind(descendants_df, row_df)
}

# Convert columns to character type
descendants_df[] <- lapply(descendants_df, as.character)

# Add the GBD_TERM column
gbd_terms <- Second_part_GBD$`GBD term`
descendants_df$GBD_TERM <- gbd_terms[match(descendants_df$MAPPED_TRAIT_URI, formatted_identifiers)]

# Reorder columns to place GBD_TERM after MAPPED_TRAIT_URI
if ("GBD_TERM" %in% colnames(descendants_df)) {
  descendants_df <- descendants_df %>% select(MAPPED_TRAIT_URI, GBD_TERM, everything())
}

# Revert the identifiers to their original format
descendants_df <- descendants_df %>% mutate(across(c(MAPPED_TRAIT_URI, starts_with("Descendant_")), ~sapply(., revert_identifier)))

# Ensure identifiers are formatted correctly in attention
attention <- attention %>% mutate(MAPPED_TRAIT_URI = format_identifier(MAPPED_TRAIT_URI))

# Remove duplicates in both data frames
attention <- attention %>% distinct(MAPPED_TRAIT_URI, .keep_all = TRUE)
descendants_df <- descendants_df %>% distinct(MAPPED_TRAIT_URI, .keep_all = TRUE)

# Reshape descendants_df to long format
descendants_long <- descendants_df %>%
  pivot_longer(cols = starts_with("Descendant_"), names_to = "Descendant", values_to = "Descendant_URI") %>%
  filter(Descendant_URI != "") %>%
  select(MAPPED_TRAIT_URI, GBD_TERM, Descendant_URI)

# Clean and standardize Descendant_URI
descendants_long$Descendant_URI <- tolower(trimws(descendants_long$Descendant_URI))
descendants_long$Descendant_URI <- gsub("\\s+", "", descendants_long$Descendant_URI)
descendants_long$Descendant_URI <- gsub(" ", "", descendants_long$Descendant_URI)
descendants_long$Descendant_URI <- gsub("[:\\s]", "", descendants_long$Descendant_URI)
descendants_long$Descendant_URI <- gsub("\"", "", descendants_long$Descendant_URI)
descendants_long$Descendant_URI <- gsub("_", ":", descendants_long$Descendant_URI)
descendants_long$Descendant_URI <- str_squish(descendants_long$Descendant_URI)
descendants_long <- descendants_long %>% distinct(Descendant_URI, .keep_all = TRUE)

# Clean and standardize MAPPED_TRAIT_URI in attention
attention$MAPPED_TRAIT_URI <- clean_mapped_trait_uri(attention$MAPPED_TRAIT_URI)

# Identify matched entries
matched_entries <- attention %>%
  filter(MAPPED_TRAIT_URI %in% descendants_long$Descendant_URI) %>%
  left_join(descendants_long %>% select(Descendant_URI, GBD_TERM), by = c("MAPPED_TRAIT_URI" = "Descendant_URI"))


 # Group by GBD term and PUBMEDID and summarize within these groups and drop na 
 matched_entries_unique <- matched_entries %>% 
    drop_na() %>%
    group_by(GBD_TERM, PUBMEDID) %>%
    summarise(
      total_attention_score_per_pubmed = first(n, na.rm = TRUE),
      weighted_n = first(weighted_n),
      nhits = first(nhits),
      weighted_nhits = first(weighted_nhits),
      weighted_attention_score_impact_factor = first(weighted_attention_score_impact_factor),
      unique_pubmed_ids = n_distinct(PUBMEDID),
      MAPPED_TRAIT_URI = first(MAPPED_TRAIT_URI),
      DISEASE_TRAIT = first(DISEASE_TRAIT),
      .groups = "drop"
    )


     # Now summarize these attention scores across the GBD term and sum the total_attention_score_per_pubmed 
  unique_matched_gbd_terms_Second_part_GBD <- matched_entries_unique %>%
    group_by(GBD_TERM) %>%
    summarise(
      total_attention_score = sum(total_attention_score_per_pubmed, na.rm = TRUE),
      weighted_n = first(weighted_n),
      nhits = first(nhits),
      weighted_nhits = first(weighted_nhits),
      weighted_attention_score_impact_factor = first(weighted_attention_score_impact_factor),
      unique_pubmed_ids = n_distinct(PUBMEDID),
      MAPPED_TRAIT_URI = first(MAPPED_TRAIT_URI),
      DISEASE_TRAIT = first(DISEASE_TRAIT),
      .groups = "drop"
    )
  
# Identify unmatched GBD entries based on GBD_TERM in both datasets
# First, extract unique GBD terms from both datasets
matched_gbd_terms <- matched_entries$GBD_TERM %>% unique()
unmatched_gbd_terms <- Second_part_GBD$`GBD term` %>% unique()

# Identify unmatched GBD terms
unique_unmatched_gbd_terms <- setdiff(unmatched_gbd_terms, matched_gbd_terms)

# Filter the Second_part_GBD dataframe to get rows with these unique unmatched GBD terms
unique_unmatched_gbd_terms_Second_part_GBD <- Second_part_GBD %>%
  filter(`GBD term` %in% unique_unmatched_gbd_terms)

# Save the data frames to Excel files
#write.xlsx(unique_unmatched_gbd_terms_Second_part_GBD, unique_unmatched_gbd_terms_Second_part_GBD.xlsx")
#write.xlsx(unique_matched_gbd_terms_Second_part_GBD, "unique_matched_gbd_terms_Second_part_GBD.xlsx")

```

Combined matched term and unmatched terms in one data set: 
unique_matched_gbd_terms_First_part_GBD & unique_matched_gbd_terms_Second_part_GBD (140)
unique_unmatched_gbd_terms_First_part_GBD & unique_unmatched_gbd_terms_Second_part_GBD (237)

```{r}

# List of columns to keep for unmatched terms
columns_unmatched <- c("GBD term", "EFO_number", "MAPPED_TRAIT_URI")

# Select only the necessary columns from the first unmatched data frame
unique_unmatched_gbd_terms_First_part_GBD <- unique_unmatched_gbd_terms_First_part_GBD[, columns_unmatched]

# Add missing columns to the second unmatched data frame
if (!"EFO_number" %in% names(unique_unmatched_gbd_terms_Second_part_GBD)) {
  unique_unmatched_gbd_terms_Second_part_GBD$EFO_number <- NA
}

# Rename the appropriate column in the second unmatched data frame
names(unique_unmatched_gbd_terms_Second_part_GBD)[which(names(unique_unmatched_gbd_terms_Second_part_GBD) == "MAPPED_TRAIT_URI...2")] <- "MAPPED_TRAIT_URI"

# Select and order the columns to match the first unmatched data frame
unique_unmatched_gbd_terms_Second_part_GBD <- unique_unmatched_gbd_terms_Second_part_GBD[, columns_unmatched]

# Combine the unmatched terms
combined_unmatched_terms <- rbind(unique_unmatched_gbd_terms_First_part_GBD, unique_unmatched_gbd_terms_Second_part_GBD)

# Display the results
print("Combined Unmatched Terms:")
print(combined_unmatched_terms)


# List of columns to keep for matched terms
columns_matched <- c("MAPPED_TRAIT_URI", "total_attention_score", "weighted_n", "nhits", "weighted_nhits", "weighted_attention_score_impact_factor", "unique_pubmed_ids", "GBD term", "DISEASE_TRAIT")

# Select only the necessary columns from the first matched data frame
unique_matched_gbd_terms_First_part_GBD <- unique_matched_gbd_terms_First_part_GBD[, columns_matched]

# Rename the column in the second matched data frame
names(unique_matched_gbd_terms_Second_part_GBD)[which(names(unique_matched_gbd_terms_Second_part_GBD) == "GBD_TERM")] <- "GBD term"

# Ensure all columns are present in the second matched data frame, filling with NA where necessary
for (col in columns_matched) {
  if (!col %in% names(unique_matched_gbd_terms_Second_part_GBD)) {
    unique_matched_gbd_terms_Second_part_GBD[[col]] <- NA
  }
}

# Select and order the columns to match the first matched data frame
unique_matched_gbd_terms_Second_part_GBD <- unique_matched_gbd_terms_Second_part_GBD[, columns_matched]

# Combine the matched terms
combined_matched_terms <- rbind(unique_matched_gbd_terms_First_part_GBD, unique_matched_gbd_terms_Second_part_GBD)

# Display the results
print("Combined Matched Terms:")
print(combined_matched_terms)

# Save the data frames to Excel files
#write.xlsx(combined_matched_terms, combined_matched_terms.xlsx")
#write.xlsx(combined_unmatched_terms, "combined_unmatched_terms.xlsx")


```

Here we could follow a partial matching approach to handle unmatched GBD terms using "str_detect" function to detect for matching but with different identifiers or GBD present within a string of traits in GWAS.

*This approach is helpful because sometimes we may miss EFO terms in the mapping process during manual mapping i.e there are many EFO terms for a particular GBD, this will be capture using str_detect" function.

To ensure consistent formatting and standardize of DISEASE/TRAIT in unmatched GBD terms and unmatched attention (GWAS) ie., both lower case for comparison, we will create new columns in both data sets 


```{r}
# Create a new column for disease names to facilitate partial matching in attention
attention <- attention %>%
  mutate(MAPPED_TRAIT_NAME = tolower(trimws(gsub("[^a-zA-Z0-9 ]", "", DISEASE_TRAIT))))

# Create a new column for disease names to facilitate partial matching in combined_unmatched_terms
combined_unmatched_terms <- combined_unmatched_terms%>%
  mutate(GBD_TERM_NAME = tolower(trimws(gsub("[^a-zA-Z0-9 ]", "", combined_unmatched_terms$`GBD term`))))


# Perform partial matching between MAPPED_TRAIT_NAME in attention and GBD_TERM_NAME in combined_unmatched_terms with the rest of the information 

partial_matches <- combined_unmatched_terms %>%
  cross_join(attention) %>%
  filter(str_detect(MAPPED_TRAIT_NAME, GBD_TERM_NAME)) %>%
  select(
    GBD_TERM_NAME, 
    MAPPED_TRAIT_NAME, 
    GBD_TRAIT_URI = MAPPED_TRAIT_URI.x, 
    MAPPED_TRAIT_URI = MAPPED_TRAIT_URI.y,
    n, 
    weighted_n, 
    nhits, 
    weighted_nhits, 
    weighted_attention_score_impact_factor,
    PUBMEDID,
    `GBD term`
  ) %>%
  distinct()



# Group by GBD term and PUBMEDID and summarize within these groups 
 partial_matches_filtered <- partial_matches %>%
    group_by(`GBD term`, PUBMEDID) %>%
    summarise(
      total_attention_score_per_pubmed = first(n, na.rm = TRUE),
      weighted_n = first(weighted_n),
      nhits = first(nhits),
      weighted_nhits = first(weighted_nhits),
      weighted_attention_score_impact_factor = first(weighted_attention_score_impact_factor),
      unique_pubmed_ids = n_distinct(PUBMEDID),
      MAPPED_TRAIT_URI = first(MAPPED_TRAIT_URI),
      GBD_TRAIT_URI = first(GBD_TRAIT_URI),
      MAPPED_TRAIT_NAME = first(MAPPED_TRAIT_NAME),
      .groups = "drop"
    )
 
     # Now summarize these attention scores across the GBD term and sum the total_attention_score_per_pubmed 
  partial_matches_unique <- partial_matches_filtered %>%
    group_by(`GBD term`) %>%
    summarise(
      total_attention_score = sum(total_attention_score_per_pubmed, na.rm = TRUE),
      weighted_n = first(weighted_n),
      nhits = first(nhits),
      weighted_nhits = first(weighted_nhits),
      weighted_attention_score_impact_factor = first(weighted_attention_score_impact_factor),
      unique_pubmed_ids = n_distinct(PUBMEDID),
      MAPPED_TRAIT_URI = first(MAPPED_TRAIT_URI),
      GBD_TRAIT_URI = first(GBD_TRAIT_URI),
      MAPPED_TRAIT_NAME = first(MAPPED_TRAIT_NAME),
      .groups = "drop"
    )



```


Identify the GBD terms that are not mapped via identifiers or trait with GWAS to combine them
in one data set. Now we will have final_matched_gbd and final_unmatched_gbd. 

```{r}


# Identify the rows that were matched partially
matched_partial_gbd <- partial_matches_unique %>% 
  select(GBD_TRAIT_URI, `GBD term`) %>% 
  distinct() 

# Get final unmatched rows
final_unmatched_gbd <- anti_join(combined_unmatched_terms, matched_partial_gbd, by = "GBD term")


# Check for duplicates in final_unmatched_gbd
duplicates_final_unmatched_gbd <- final_unmatched_gbd[duplicated(final_unmatched_gbd$`GBD term`), ]
if (nrow(duplicates_final_unmatched_gbd) > 0) {
  print("Duplicates found in final_unmatched_gbd:")
  print(duplicates_final_unmatched_gbd)
} else {
  print("No duplicates found in final_unmatched_gbd.")
}


# final matched_gbd (combine those identified via Identifier and Partially)
final_matched_gbd <- bind_rows(
  combined_matched_terms %>% mutate(Match_Type = "Identifier"),
  partial_matches_unique %>% mutate(Match_Type = "Partial")
)

# Check for duplicates in final matched_gbd
duplicates_final_matched_gbd <- final_matched_gbd[duplicated(final_matched_gbd$`GBD term`), ]
if (nrow(duplicates_final_matched_gbd) > 0) {
  print("Duplicates found in final_matched_gbd:")
  print(duplicates_final_matched_gbd)
} else {
  print("No duplicates found in final_matched_gbd")
}


# Save the results to a CSV file
#write.csv(final_unmatched_gbd, "final_unmatched_gbd.csv", row.names = FALSE)
#write.csv(final_matched_gbd, "final_matched_gbd.csv", row.names = FALSE)


```


If there are no matches for a GBD trait then the attention score = 0. also, we ill combine matced and unmatched GBD terms in one data set (combined dataset). It has 377 GBD terms, as we started. 

```{r}


#  Add 'total_attention_score' column with value 0 to final_unmatched_gbd
final_unmatched_gbd <- final_unmatched_gbd %>%
  mutate(total_attention_score = 0)

#  Ensure both datasets have the same columns
# Here, we ensure that final_unmatched_gbd has all necessary columns
required_columns <- names(final_matched_gbd)

# Add any missing columns to final_unmatched_gbd and set their values to NA
for (col in required_columns) {
  if (!(col %in% names(final_unmatched_gbd))) {
    final_unmatched_gbd[[col]] <- NA
  }
}

# Combine the datasets
combined_dataset <- bind_rows(final_matched_gbd, final_unmatched_gbd)

combined_dataset <- combined_dataset %>%
  distinct(`GBD term`, .keep_all = TRUE)


# Save the combined dataset to a CSV file
#write.csv(combined_dataset, "combined_dataset.csv", row.names = FALSE)

```


 load the GBD DALY data, verify any missing values, and inspect the structure of the dataset
 
```{r}

# Load the GBD DALY data
gbd_daly_data <- read_csv('/Users/rayanaloliet/Desktop/GWAS Catalog/IHME-GBD_2021_DATA-120ebfcd-1.csv')

# Filter out rows with NA values in the key columns
gbd_daly_data <- gbd_daly_data %>%
  filter(!is.na(`cause_name`), !is.na(val))

# Verify that there are no NA values left
sum(is.na(gbd_daly_data$`cause_name`))
sum(is.na(gbd_daly_data$val))
sum(is.na(gbd_daly_data$measure_name))

 
# Inspect the structure of the GBD DALY data
str(gbd_daly_data)




```


Assessing the relationship between GWAS attention and global need, using an inequality measure.This will be done via filtering the data by Number of DALY and then merging by GBD term.

```{r}
#2019 
#gbd_daly_data <- gbd_daly_data %>%
 # filter(metric_name == "Number" & measure_name == "DALYs (Disability-Adjusted Life Years)") #%>%
 # rename(`GBD term`= cause_name, DALY = val)
#2021
gbd_daly_data <- gbd_daly_data %>%
  filter(metric_name == "Number") %>%
  rename(`GBD term`= cause_name, DALY = val)


# Merge the DALY values into the combined dataset
combined_dataset <- combined_dataset %>%
  left_join(gbd_daly_data, by = c("GBD term" = "GBD term"))

# Remove specified columns
combined_dataset <- combined_dataset %>%
  select(-c(DISEASE_TRAIT, GBD_TRAIT_URI, MAPPED_TRAIT_NAME, EFO_number, GBD_TERM_NAME))

# Verify the merge
head(combined_dataset)

# Ensure the columns are numeric as it gives error massage 
combined_dataset$total_attention_score <- as.numeric(combined_dataset$total_attention_score)
combined_dataset$DALY <- as.numeric(combined_dataset$DALY)

# Check for any NA values and handle them
combined_dataset[is.na(combined_dataset$DALY), 'DALY'] <- 0

# Verify the data
summary(combined_dataset$total_attention_score)
summary(combined_dataset$DALY)


```



Try to identify any difference or duplicates between combined dataset and the data set that we started with in the manual mapping.  
 

```{r}
unmatched_gbd_terms <- anti_join(gbd_daly_data, combined_dataset, by = c("GBD term" = "GBD term"))

unmatched_gbd_terms <- unmatched_gbd_terms %>%
  distinct(`GBD term`, .keep_all = TRUE)

# Get the unique GBD terms in each dataset
combined_gbd_terms <- unique(combined_dataset$`GBD term`)
gbd_gbd_terms <- unique(gbd_daly_data$`GBD term`)

# Identify terms in combined_dataset not in gbd_daly_data
terms_only_in_combined <- setdiff(combined_gbd_terms, gbd_gbd_terms)

# Identify terms in gbd_daly_data not in combined_dataset
terms_only_in_gbd_daly <- setdiff(gbd_gbd_terms, combined_gbd_terms)

# Get the lengths of the unique terms
len_combined <- length(terms_only_in_combined)
len_gbd_daly <- length(terms_only_in_gbd_daly)

# Determine the maximum length to ensure both columns in the data frame have the same number of rows
max_length <- max(len_combined, len_gbd_daly)

# Combine the differences into a data frame for better visualization
diff_2 <- data.frame(
  Terms_Only_in_Combined_Dataset = c(terms_only_in_combined, rep(NA, max_length - len_combined)),
  Terms_Only_in_GBD_Daly_Dataset = c(terms_only_in_gbd_daly, rep(NA, max_length - len_gbd_daly))
)

# the differences
(diff_2)

#write.csv(diff_2, "diff_2.csv", row.names = FALSE)


```

Visualizing the distribution of total_attention_score 

```{r}

# Visualize the distribution of 'total_attention_score'
ggplot(combined_dataset, aes(x = total_attention_score)) +
  geom_histogram(binwidth = 500, fill = 'blue', color = 'black', alpha = 0.7) +
  labs(title = "Histogram of Total Attention Scores", x = "total_attention_score", y = "Frequency") +
  xlim(c(0, 35000))  


# Adding 1 to avoid log(0) which is undefined, for better visualization 
combined_dataset$total_attention_score_log <- log10(combined_dataset$total_attention_score + 1)

# Visualize the distribution with log transformation
ggplot(combined_dataset, aes(x = total_attention_score_log)) +
  geom_histogram(binwidth = 0.1, fill = 'blue', color = 'black', alpha = 0.7) +
  labs(title = "Histogram of Log-Transformed Total Attention Scores (log(n + 1))", 
       x = "Log10(total attention Score + 1)", 
       y = "Frequency") +
  scale_x_continuous(breaks = seq(0, log10(max(combined_dataset$total_attention_score + 1)), by = 1))


# Boxplot for 'total_attention_score'
ggplot(combined_dataset, aes(y = total_attention_score)) +
  geom_boxplot(fill = 'blue', color = 'black') +
  labs(title = "Boxplot of Total Attention Scores (n)", y = "Total Attention Score (n)")

```



Visualizing the distribution of daly and total_attention_score

```{r}


# Create the histogram for DALY distribution
ggplot(combined_dataset, aes(x = DALY)) +
  geom_histogram(binwidth = 10000, aes(fill = after_stat(count)), color = 'black', alpha = 0.7) +
  scale_fill_gradient(low = "red", high = "darkred") +
  labs(title = "Histogram of DALY", x = "DALY", y = "Frequency") +
  theme_minimal()


# Boxplot for 'DALY'
ggplot(combined_dataset, aes(y = DALY)) +
  geom_boxplot(fill = 'red', color = 'black') +
  labs(title = "Boxplot of DALY", y = "DALY")


```

Visualizing the distribution of daly and total attention score

```{r}
# Scatter plot to visualize the relationship between 'total attention score' and 'DALY'
ggplot(combined_dataset, aes(x = DALY, y = total_attention_score)) +
  geom_point(color = 'purple') +
  labs(title = "Scatter Plot of DALY vs Total Attention Scores (n)", x = "DALY", y = "Total Attention Score (n)")

# Scatter plot to visualize the relationship between 'weighted_attention_score_impact_factor' and 'DALY'
filtered_dataset <- combined_dataset %>%
  filter(!is.na(DALY) & !is.na(weighted_attention_score_impact_factor))

ggplot(filtered_dataset, aes(x = DALY, y = weighted_attention_score_impact_factor)) +
  geom_point(color = 'blue') +
  labs(title = "Scatter Plot of DALY vs Weighted Attention Score Impact Factor", 
       x = "DALY", 
       y = "Weighted Attention Score Impact Factor")
```




filtering the data and remove those with total_attention_score= zero to have a better understating of the other GBD term and visualize the filtered data

```{r}
# check the summary of the total_attention_score column 
summary(combined_dataset$total_attention_score)

# Check the count of rows where total_attention_score is zero
sum(combined_dataset$total_attention_score == 0)


# Filter out rows with zero total attention scores
filtered_dataset <- combined_dataset %>% filter(total_attention_score > 0)

# Check the summary of the filtered data
summary(filtered_dataset$total_attention_score)


# Summary of 'total_attention_score' in the filtered dataset
summary(filtered_dataset$total_attention_score)



```

Visualizing the distribution of total attention score after filtering zero values 

```{r}
# Histogram of Attention Scores (n)
ggplot(filtered_dataset, aes(x = total_attention_score)) +
  geom_histogram(binwidth = 1, fill = 'blue', color = 'black', alpha = 0.7) +
  labs(title = "Histogram of Attention Scores (n) - Filtered", x = "Attention Score (n)", y = "Frequency")


```

Investigate Specific Diseases: Identify specific diseases with high DALY but low attention scores

a threshold for low attention: Here, I have used n <= 1 as the threshold for low attention. 

```{r}
# Filter for diseases with high DALY but low attention scores
threshold_attention = 1  # Define what is considered low attention
high_daly_low_attention <- combined_dataset %>%
  filter(DALY > median(DALY) & total_attention_score <= threshold_attention) %>%
  arrange(desc(DALY))

# View the top diseases with high DALY and low attention
(head(high_daly_low_attention, 20))

# Save the filtered data to a CSV file
#write.csv(high_daly_low_attention, "high_daly_low_attention.csv", row.names = FALSE)



```
identify the trait hat has been investigated the most based on total_attention_score

```{r}

summary(combined_dataset$total_attention_score)
frequency <- table(combined_dataset$total_attention_score)


specific_value <- 33796

# Filter the dataset for the specific total_attention_score_per_pubmed value
filtered_rows <- combined_dataset %>%
  filter(total_attention_score == specific_value)

# Extract unique GBD terms
unique_gbd_terms <- unique(filtered_rows$`GBD term`)

# Print the unique GBD terms
print(unique_gbd_terms)
```

identify top 50 GBD terms investigated by GWAS

```{r}
# Sort the data frame by total_attention_score in descending order
sorted_data <- combined_dataset[order(combined_dataset$total_attention_score, decreasing = TRUE), ]

# Remove duplicates based on total_attention_score and GBD term combination
unique_data <- sorted_data[!duplicated(sorted_data[, c("total_attention_score", "GBD term")]), ]

# Select the top 50 unique values
top_50_scores <- head(unique_data, 50)

# Display the top 50 scores
print(top_50_scores)


# Threshold value
threshold_value <- 198

# Filter the dataset for total_attention_score_per_pubmed values greater than the threshold
filtered_rows <- combined_dataset %>%
  filter(total_attention_score > threshold_value)

# Extract unique GBD terms
unique_gbd_terms <- unique(filtered_rows$`GBD term`)

# Print the unique GBD terms
print(unique_gbd_terms)

# save
#write.csv(unique_gbd_terms, file = "unique_gbd_terms.csv", row.names = FALSE)

```
identify the rank of DALY for the top 50 GBD terms investigated by GWAS

```{r}

# Sort the data frame by total_attention_score in descending order and remove duplicates based on total_attention_score and GBD term combination
unique_data <- combined_dataset %>%
  arrange(desc(total_attention_score)) %>%
  distinct(total_attention_score, `GBD term`, .keep_all = TRUE)

# Rank the DALY values based on the deduplicated dataset
unique_data <- unique_data %>%
  mutate(DALY_rank = rank(DALY))

# Threshold value
threshold_value <- 198

# Filter the dataset for total_attention_score values greater than the threshold
filtered_rows <- unique_data %>%
  filter(total_attention_score > threshold_value)

# Summarize the data to get frequency and list of unique total_attention_score and DALY for each GBD term
summary_table <- filtered_rows %>%
  group_by(`GBD term`) %>%
  summarise(
    total_attention_score = unique(total_attention_score),
    DALY = unique(DALY),
    DALY_rank = unique(DALY_rank),
    .groups = "drop"
  )

# View the summary table
print(summary_table)

```


Developing the concentration Index and Lorenz Curve; using Discrete Sum Method

```{r}
location_names <- unique(combined_dataset$location_name)
valid_data_threshold <- 10  # Define a threshold for valid data points

concentration_results <- data.frame(
  location_name = character(),
  concentration_index = numeric(),
  ci_lower = numeric(),
  ci_upper = numeric(),
  total_points = numeric(),
  valid_points = numeric(),
  stringsAsFactors = FALSE
)

locations_failed <- c()
locations_passed <- c()

# Define the concentration index calculation function
concentration_index_function <- function(data, indices) {
  # Resample data
  resampled_data <- data[indices, ]
  
  # Sort the data by attention scores
  sorted_indices <- order(resampled_data$attention_scores)
  sorted_attention_scores <- resampled_data$attention_scores[sorted_indices]
  sorted_daly_values <- resampled_data$daly_values[sorted_indices]
  
  # Calculate cumulative proportions
  cumulative_attention_scores <- cumsum(sorted_attention_scores) / sum(sorted_attention_scores)
  cumulative_daly_values <- cumsum(sorted_daly_values) / sum(sorted_daly_values)
  
  # Calculate the concentration index
  concentration_index <- 2 * sum(cumulative_daly_values * sorted_attention_scores) / sum(sorted_attention_scores) - 1
  
  return(concentration_index)
}

for (location in location_names) {
  cat("Processing location:", location, "\n")
  
  subset_data <- combined_dataset %>%
    filter(location_name == !!location)
  
  total_attention_scores <- subset_data$total_attention_score
  daly_values <- subset_data$DALY
  
  # Remove rows with missing values
  valid_data <- subset_data %>%
    filter(!is.na(total_attention_scores) & !is.na(daly_values))
  
  total_attention_scores <- valid_data$total_attention_score
  daly_values <- valid_data$DALY
  
  # Check for non-positive values
  valid_points <- nrow(valid_data)
  total_points <- nrow(subset_data)
  
  if (valid_points < valid_data_threshold) {
    cat("Skipping location:", location, "- not enough valid data points.\n")
    cat("Total points:", total_points, "- Valid points:", valid_points, "\n")
    concentration_results <- concentration_results %>%
      add_row(
        location_name = location,
        concentration_index = NA,
        ci_lower = NA,
        ci_upper = NA,
        total_points = total_points,
        valid_points = valid_points
      )
    locations_failed <- c(locations_failed, location)
    next
  }
  
  data <- data.frame(attention_scores = total_attention_scores, daly_values = daly_values)
  
  # Check for zero variance in data
  if (var(total_attention_scores) == 0 || var(daly_values) == 0) {
    cat("Zero variance in data for location:", location, "\n")
    concentration_results <- concentration_results %>%
      add_row(
        location_name = location,
        concentration_index = NA,
        ci_lower = NA,
        ci_upper = NA,
        total_points = total_points,
        valid_points = valid_points
      )
    locations_failed <- c(locations_failed, location)
    next
  }
  
  # Print the data distribution for debugging
  cat("Total Attention Scores - Min:", min(total_attention_scores), "Max:", max(total_attention_scores), "Mean:", mean(total_attention_scores), "\n")
  cat("DALY Values - Min:", min(daly_values), "Max:", max(daly_values), "Mean:", mean(daly_values), "\n")
  
  set.seed(123)  # for reproducibility
  bootstrap_results <- tryCatch({
    boot(data, concentration_index_function, R = 1000)  # Increase the number of replications if needed
  }, error = function(e) {
    cat("Error in bootstrapping for location:", location, "\n")
    return(NULL)
  })
  
  if (is.null(bootstrap_results)) {
    cat("Skipping location:", location, "due to bootstrapping error.\n")
    concentration_results <- concentration_results %>%
      add_row(
        location_name = location,
        concentration_index = NA,
        ci_lower = NA,
        ci_upper = NA,
        total_points = total_points,
        valid_points = valid_points
      )
    locations_failed <- c(locations_failed, location)
    next
  }
  
  bootstrap_ci <- tryCatch({
    boot.ci(bootstrap_results, type = "perc")
  }, error = function(e) {
    cat("Error in confidence interval calculation for location:", location, "\n")
    return(NULL)
  })
  
  if (is.null(bootstrap_ci)) {
    cat("Skipping location:", location, "due to confidence interval calculation error.\n")
    concentration_results <- concentration_results %>%
      add_row(
        location_name = location,
        concentration_index = NA,
        ci_lower = NA,
        ci_upper = NA,
        total_points = total_points,
        valid_points = valid_points
      )
    locations_failed <- c(locations_failed, location)
    next
  }
  
  concentration_index <- concentration_index_function(data, 1:nrow(data))
  
  concentration_results <- concentration_results %>%
    add_row(
      location_name = location,
      concentration_index = concentration_index,
      ci_lower = bootstrap_ci$percent[4],
      ci_upper = bootstrap_ci$percent[5],
      total_points = total_points,
      valid_points = valid_points
    )
  
  # Calculate cumulative proportions for plotting the concentration curve
  sorted_indices <- order(total_attention_scores)
  cumulative_attention_scores <- cumsum(total_attention_scores[sorted_indices]) / sum(total_attention_scores)
  cumulative_daly_values <- cumsum(daly_values[sorted_indices]) / sum(daly_values)
  
  # Plot the concentration curve
  plot(cumulative_daly_values, cumulative_attention_scores, type="l", col="blue", lwd=2, 
       xlab="Cumulative Proportion of DALY Values", ylab="Cumulative Proportion of Attention Scores",
       main=paste("Concentration Curve -", location))
  abline(0, 1, col="red", lwd=2, lty=2) # Line of equality
  
  cat("Displayed plot for location:", location, "\n")
  locations_passed <- c(locations_passed, location)
}

# Print summary of results
print(concentration_results)
cat("Locations passed:", paste(locations_passed, collapse = ", "), "\n")
cat("Locations failed:", paste(locations_failed, collapse = ", "), "\n")

# Combined concentration curve for all data
all_attention_scores <- combined_dataset$total_attention_score
all_daly_values <- combined_dataset$DALY

# Remove rows with missing values
valid_data_all <- combined_dataset %>%
  filter(!is.na(total_attention_score) & !is.na(DALY))

all_attention_scores <- valid_data_all$total_attention_score
all_daly_values <- valid_data_all$DALY

# Prepare data for the combined plot
all_data <- data.frame(attention_scores = all_attention_scores, daly_values = all_daly_values)

# Calculate the concentration index and bootstrap CI for all data
all_concentration_index <- concentration_index_function(all_data, 1:nrow(all_data))

set.seed(123)  # for reproducibility
all_bootstrap_results <- boot(all_data, concentration_index_function, R = 1000)

# Calculate the 95% confidence interval for the combined data
all_bootstrap_ci <- boot.ci(all_bootstrap_results, type = "perc")

# Calculate cumulative proportions for plotting the combined concentration curve
sorted_indices_all <- order(all_attention_scores)
cumulative_attention_scores_all <- cumsum(all_attention_scores[sorted_indices_all]) / sum(all_attention_scores)
cumulative_daly_values_all <- cumsum(all_daly_values[sorted_indices_all]) / sum(all_daly_values)

# Plot the combined concentration curve
plot(cumulative_daly_values_all, cumulative_attention_scores_all, type="l", col="blue", lwd=2, 
     xlab="Cumulative Proportion of DALY Values", ylab="Cumulative Proportion of Attention Scores",
     main="Combined Concentration Curve")
abline(0, 1, col="red", lwd=2, lty=2) # Line of equality

# Print the combined concentration index and its confidence interval
cat("Combined Concentration Index:", all_concentration_index, "\n")
cat("95% Confidence Interval:", all_bootstrap_ci$percent[4], "-", all_bootstrap_ci$percent[5], "\n")
```


Developing the concentration Index and Lorenz Curve; using Numerical Integration Method

```{r}
location_names <- unique(combined_dataset$location_name)
valid_data_threshold <- 10  # Define a threshold for valid data points

concentration_results <- data.frame(
  location_name = character(),
  concentration_index = numeric(),
  ci_lower = numeric(),
  ci_upper = numeric(),
  total_points = numeric(),
  valid_points = numeric(),
  stringsAsFactors = FALSE
)

locations_failed <- c()
locations_passed <- c()

# Define the concentration index calculation function
concentration_index_function <- function(data, indices) {
  resampled_data <- data[indices, ]
  resampled_curve <- Lorenz.curve(y = resampled_data$total_attention_score, x = resampled_data$DALY, graph = FALSE, na.rm = TRUE)
  return(tryCatch({
    integrate(resampled_curve, lower = 0, upper = 1, subdivisions = 10000)$value * 2 - 1
  }, error = function(e) {
    return(NA)
  }))
}

for (location in location_names) {
  cat("Processing location:", location, "\n")
  
  subset_data <- combined_dataset %>%
    filter(location_name == !!location)
  
  total_attention_scores <- subset_data$total_attention_score
  daly_values <- subset_data$DALY
  
  # Remove rows with missing values
  valid_data <- subset_data %>%
    filter(!is.na(total_attention_scores) & !is.na(daly_values))
  
  total_attention_scores <- valid_data$total_attention_score
  daly_values <- valid_data$DALY
  
  # Check for valid data points
  valid_points <- nrow(valid_data)
  total_points <- nrow(subset_data)
  
  if (valid_points < valid_data_threshold) {
    cat("Skipping location:", location, "- not enough valid data points.\n")
    cat("Total points:", total_points, "- Valid points:", valid_points, "\n")
    concentration_results <- concentration_results %>%
      add_row(
        location_name = location,
        concentration_index = NA,
        ci_lower = NA,
        ci_upper = NA,
        total_points = total_points,
        valid_points = valid_points
      )
    locations_failed <- c(locations_failed, location)
    next
  }
  
  # Compute the concentration curve and index using Lorenz.curve
  concentration_curve <- Lorenz.curve(y = total_attention_scores, x = daly_values, graph = FALSE, na.rm = TRUE)
  
  # Calculate the concentration index with increased maximum subdivisions
  concentration_index <- tryCatch({
    integrate(concentration_curve, lower = 0, upper = 1, subdivisions = 10000)$value * 2 - 1
  }, error = function(e) {
    cat("Error in calculating concentration index for location:", location, "\n")
    return(NA)
  })
  
  if (!is.na(concentration_index)) {
    # Perform bootstrapping to calculate 95% confidence intervals
    data <- data.frame(total_attention_score = total_attention_scores, DALY = daly_values)
    set.seed(123)  # for reproducibility
    bootstrap_results <- boot(data, concentration_index_function, R = 1000)
    bootstrap_ci <- tryCatch({
      boot.ci(bootstrap_results, type = "perc")
    }, error = function(e) {
      cat("Error in calculating confidence interval for location:", location, "\n")
      return(NULL)
    })
    
    if (!is.null(bootstrap_ci)) {
      ci_lower <- bootstrap_ci$percent[4]
      ci_upper <- bootstrap_ci$percent[5]
    } else {
      ci_lower <- NA
      ci_upper <- NA
    }
  } else {
    ci_lower <- NA
    ci_upper <- NA
  }
  
  concentration_results <- concentration_results %>%
    add_row(
      location_name = location,
      concentration_index = concentration_index,
      ci_lower = ci_lower,
      ci_upper = ci_upper,
      total_points = total_points,
      valid_points = valid_points
    )
  
  # Plot the concentration curve with the location name
  plot(concentration_curve, main = paste("Concentration Curve -", location), col = "blue", lwd = 2, 
       xlab = "Cumulative Proportion of DALY Values", ylab = "Cumulative Proportion of Attention Scores")
  abline(0, 1, col = "red", lwd = 2, lty = 2)  # Line of equality
  
  cat("Displayed plot for location:", location, "\n")
  locations_passed <- c(locations_passed, location)
}

# Print summary of results
print(concentration_results)
cat("Locations passed:", paste(locations_passed, collapse = ", "), "\n")
cat("Locations failed:", paste(locations_failed, collapse = ", "), "\n")

# Combined concentration curve for all data
all_attention_scores <- combined_dataset$total_attention_score
all_daly_values <- combined_dataset$DALY

# Remove rows with missing values
valid_data_all <- combined_dataset %>%
  filter(!is.na(total_attention_score) & !is.na(DALY))

all_attention_scores <- valid_data_all$total_attention_score
all_daly_values <- valid_data_all$DALY

# Compute the combined concentration curve and index using Lorenz.curve
combined_concentration_curve <- Lorenz.curve(y = all_attention_scores, x = all_daly_values, graph = FALSE, na.rm = TRUE)

# Plot the combined concentration curve
plot(combined_concentration_curve, main = "Combined Concentration Curve", col = "blue", lwd = 2, 
     xlab = "Cumulative Proportion of DALY Values", ylab = "Cumulative Proportion of Attention Scores")
abline(0, 1, col = "red", lwd = 2, lty = 2)  # Line of equality

# Calculate the combined concentration index with increased maximum subdivisions
combined_concentration_index <- tryCatch({
  integrate(combined_concentration_curve, lower = 0, upper = 1, subdivisions = 10000)$value * 2 - 1
}, error = function(e) {
  cat("Error in calculating combined concentration index\n")
  return(NA)
})

# Perform bootstrapping for combined data to calculate 95% confidence intervals
combined_data <- data.frame(total_attention_score = all_attention_scores, DALY = all_daly_values)

set.seed(123)  # for reproducibility
combined_bootstrap_results <- boot(combined_data, concentration_index_function, R = 1000)

combined_bootstrap_ci <- tryCatch({
  boot.ci(combined_bootstrap_results, type = "perc")
}, error = function(e) {
  cat("Error in calculating confidence interval for combined data\n")
  return(NULL)
})

# Print the combined concentration index and its 95% confidence interval
cat("Combined Concentration Index:", combined_concentration_index, "\n")
if (!is.null(combined_bootstrap_ci)) {
  cat("95% Confidence Interval:", combined_bootstrap_ci$percent[4], "-", combined_bootstrap_ci$percent[5], "\n")
} else {
  cat("95% Confidence Interval: Not available due to error\n")
}

# Append the combined data results
combined_entry <- data.frame(
  location_name = "Combined",
  concentration_index = combined_concentration_index,
  ci_lower = ifelse(!is.null(combined_bootstrap_ci), combined_bootstrap_ci$percent[4], NA),
  ci_upper = ifelse(!is.null(combined_bootstrap_ci), combined_bootstrap_ci$percent[5], NA),
  total_points = nrow(combined_dataset),
  valid_points = nrow(valid_data_all)
)

concentration_results <- rbind(concentration_results, combined_entry)

# Sort the DataFrame by concentration index for better visualization
concentration_results <- concentration_results %>%
  arrange(desc(concentration_index))

# Plot concentration index with 95% confidence intervals for all locations individually and combined
ggplot(concentration_results, aes(x = reorder(location_name, concentration_index), y = concentration_index)) +
  geom_point() +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper)) +
  coord_flip() +
  labs(
    title = "Concentration Index by Location",
    x = "Location",
    y = "Concentration Index"
  ) +
  theme_minimal()
```


Remove non-genetic GBD terms for those that not  possible to gain insight from GWAS studies. 

```{r}
# List of terms to exclude
terms_to_exclude <- c("All causes",
                      "Cirrhosis due to alcohol",
                      "Low back pain",
                      "Neck pain",
                      "Cocaine use disorders",
                      "Drug use disorders",
                      "Foreign body in other body part",
                      "Opioid use disorders",
                      "Poisonings",
                      "Falls",
                      "Foreign body",
                      "Conflict and terrorism",
                      "Drowning",
                      "Poisoning by carbon monoxide",
                      "Extensively drug-resistant tuberculosis",
                      "Venomous animal contact",
                      "Sexual violence",
                      "Adverse effects of medical treatment",
                      "Pulmonary aspiration and foreign body in airway",
                      "Motorcyclist road injuries",
                      "Motor vehicle road injuries",
                      "Physical violence by firearm",
                      "Physical violence by sharp object",
                      "Environmental heat and cold exposure",
                      "Police conflict and executions",
                      "Exposure to forces of nature",
                      "Animal contact",
                      "Non-venomous animal contact",
                      "Fire, heat, and hot substances",
                      "Unintentional firearm injuries",
                      "Indirect maternal deaths",
                      "Late maternal deaths",
                      "Amphetamine use disorders",
                      "Other drug use disorders",
                      "Other infectious diseases",
                      "Other transport injuries",
                      "Unintentional injuries",
                      "Other unintentional injuries",
                      "Poisoning by other means",
                      "Exposure to mechanical forces",
                      "Physical violence by other means",
                      "Injuries",
                      "Transport injuries",
                      "Road injuries",
                      "Other road injuries",
                      "Foreign body in eyes",
                      "Cannabis use disorders")

# Filter the GBD DALY data for "Number" metric and rename columns
combined_dataset <- combined_dataset %>%
  filter(!(`GBD term` %in% terms_to_exclude))  # Exclude specified terms


```

here we could go again and run the codes of the two methods above after removing those GBD terms to see the impact 
