---
title: "Example analysis of Rayan's project"
output: html_notebook
---


Read in the data

```{r}
library(dplyr)
library(data.table)
library(tidyr)
library(stringr)

# a <- fread("~/Downloads/gwas_catalog_v1.0.2.1-studies_r2024-03-11.tsv")

a <- fread("/Users/rayanaloliet/Desktop/GWAS Catalog/gwas_catalog_v1.0.2.1-studies_r2024-06-07.csv")

```



Have a look at the data

```{r}
hist(a$`ASSOCIATION COUNT`, breaks=100)
```

Sometimes there is more than one EFO term (comma separated). Split them into separate lines

```{r}

head(a, 10)

a <- a %>%
    tidyr::separate_rows(MAPPED_TRAIT_URI, sep = ", ")

head(a, 10)

```

Get the number of EFO terms per publication

```{r}
pubmed_count <- a %>% group_by(PUBMEDID) %>% 
    summarise(
        n_efo = length(unique(MAPPED_TRAIT_URI))
    ) %>% arrange(desc(n_efo))

pubmed_count
```

Add the EFO count per PUBMEDID to the data frame

```{r}
a <- left_join(a, pubmed_count, by="PUBMEDID")
str(a)
```

Step 1. Develop the attention scores

- n = number of studies with that EFO term
- weighted_n = weighted by number of EFO terms in the study in total
- nhits = number of GWAS hits reported for that EFO term
- weighted_nhits = nhits weighted by number of EFO terms in the study in total

```{r}
attention <- a %>% group_by(
    MAPPED_TRAIT_URI
) %>% 
    summarise(
        n = n(), 
        weighted_n = sum(1/n_efo),
        nhits = sum(`ASSOCIATION COUNT`),
        weighted_nhits = sum(`ASSOCIATION COUNT` / n_efo)
    )

str(attention)
```


Removing extra whitespace, cleaning the Impact factor values and ensure the values are numeric 

```{r}
# Remove extra whitespace
a$`Impact factor` <- trimws(a$`Impact factor`)

# Replace non-standard characters
a$`Impact factor` <- gsub("·", ".", a$`Impact factor`)

# Convert to numeric, forcing non-numeric values to NA (for those with no impact factors)
a$`Impact factor` <- as.numeric(a$`Impact factor`)

# Replace NA values with 0
a$`Impact factor`[is.na(a$`Impact factor`)] <- 0

# Verify the changes
str(a)
unique(a$`Impact factor`)
summary(a$`Impact factor`)


```
 Including the impact factor values after preparing and cleaning to the attention scores
 
```{r}

attention <- a %>% 
  group_by(MAPPED_TRAIT_URI) %>% 
  summarise(
    n = n(), 
    weighted_n = sum(1 / n_efo),
    nhits = sum(`ASSOCIATION COUNT`),
    weighted_nhits = sum(`ASSOCIATION COUNT` / n_efo),
    weighted_attention_score_impact_factor = sum((1 / n_efo) * `Impact factor`, na.rm = TRUE)
  )

str(attention)
str(a)

```
 
Step 2. Merging the GWAS attention scores with the GBD disease burden results

```{r}

gbd_data <- fread("/Users/rayanaloliet/Desktop/GWAS Catalog/GDB.csv")
str(gbd_data)

```

Specify columns that may contain EFO terms

```{r}

efo_columns <- c("EFO 1", "EFO 2", "EFO 3", "EFO 4", "EFO 5", "EFO 6", 
                 "EFO 7", "EFO 8", "EFO 9", "EFO 10", "EFO 11", "EFO 12",
                 "EFO 13", "EFO 14", "EFO 15", "EFO 16", "EFO 17", "EFO 18",
                 "EFO 19", "EFO 20", "EFO 21", "EFO 22", "EFO 23", "EFO 24",
                 "EFO 25", "EFO 26", "EFO 27", "EFO 28", "EFO 29", "EFO 30")
```


Reshape the GBD data to gather all relevant columns into key-value pairs and drop rows with NA EFO terms

```{r}
gbd_long <- gbd_data %>%
  pivot_longer(cols = all_of(efo_columns), names_to = "EFO_number", values_to = "MAPPED_TRAIT_URI") %>%
  drop_na(MAPPED_TRAIT_URI) 

gbd_long
```

Separate rows based on a separator (assuming comma-separated EFO terms) and Trim whitespace

```{r}
gbd_long <- gbd_long %>%
  separate_rows(MAPPED_TRAIT_URI, sep = ",") %>%
  mutate(MAPPED_TRAIT_URI = trimws(MAPPED_TRAIT_URI))  

gbd_long
```

Ensure consistent formatting, cleaning and standardize MAPPED_TRAIT_URI in gbd_long (GBD file)

*There are many functions because each time I attempted to find a matching, I was unable to do so due to the differences in the two datasets. For example, in GWAS, the identifiers have underscores, while in GBD they have colons, and so on... Therefore, the functions below are intended to unify these differences. 

```{r}

# Ensure consistent formatting
gbd_long$MAPPED_TRAIT_URI <- tolower(trimws(gbd_long$MAPPED_TRAIT_URI))

# Ensure lowercase
gbd_long$MAPPED_TRAIT_URI <- tolower(gbd_long$MAPPED_TRAIT_URI) 

# Trim leading/trailing whitespace
gbd_long$MAPPED_TRAIT_URI <- trimws(gbd_long$MAPPED_TRAIT_URI) 

# Remove all spaces
gbd_long$MAPPED_TRAIT_URI <- gsub("\\s+", "", gbd_long$MAPPED_TRAIT_URI) 

# Remove all internal spaces explicitly
gbd_long$MAPPED_TRAIT_URI <- gsub(" ", "", gbd_long$MAPPED_TRAIT_URI) 

# Convert to lowercase and remove spaces 
gbd_long$MAPPED_TRAIT_URI <- tolower(gsub("\\s+", "", gbd_long$MAPPED_TRAIT_URI)) 

# Remove colons and spaces
gbd_long$MAPPED_TRAIT_URI <- gsub("[:\\s]", "", gbd_long$MAPPED_TRAIT_URI)  

# Remove quotation marks
gbd_long$MAPPED_TRAIT_URI <- gsub("\"", "", gbd_long$MAPPED_TRAIT_URI) 

# Replace _ with : 
gbd_long$MAPPED_TRAIT_URI <- gsub("_", ":", gbd_long$MAPPED_TRAIT_URI) 

# Remove all extra spaces
gbd_long$MAPPED_TRAIT_URI <- str_squish(gbd_long$MAPPED_TRAIT_URI) 

print(head(gbd_long$MAPPED_TRAIT_URI, 20))

```


```{r}

(head(attention))
#rm(list = ls())

```




Ensure consistent formatting, cleaning and standardize MAPPED_TRAIT_URI in attention (GWAS file)

```{r}

# Function to clean and standardize MAPPED_TRAIT_URI
clean_mapped_trait_uri <- function(uri) {
  # Convert to lowercase
  uri <- tolower(uri)
  # Trim leading/trailing whitespace
  uri <- trimws(uri)
  # Remove all internal spaces explicitly
  uri <- gsub("\\s+", "", uri)
  # Remove quotation marks
  uri <- gsub("\"", "", uri)
  # Replace _ with :
  uri <- gsub("_", ":", uri)
  # Remove colons and spaces
  uri <- gsub("[:\\s]", "", uri)
  # Remove prefixes and ensure consistency
  uri <- sub(".*[:/]", "", uri)
  return(uri)
}

# Apply the function to MAPPED_TRAIT_URI column
attention$MAPPED_TRAIT_URI <- clean_mapped_trait_uri(attention$MAPPED_TRAIT_URI)

# Print the first 20 cleaned URIs
print(head(attention$MAPPED_TRAIT_URI, 20))


```


Matching GWAS to GBD via MAPPED_TRAIT_URI after cleaning 

```{r}
# Merge the data frames
merged_data <- merge(attention, gbd_long, by = "MAPPED_TRAIT_URI", all.x = TRUE)

# Display the first few rows of the merged data frame
head(merged_data)

# Identify unmatched rows in attention
unmatched_attention <- anti_join(attention, gbd_long, by = "MAPPED_TRAIT_URI")
unmatched_attention
(head(unmatched_attention, 20))


# Identify unmatched rows in gbd_long
unmatched_gbd <- anti_join(gbd_long, attention, by = "MAPPED_TRAIT_URI")
unmatched_gbd
(head(unmatched_gbd, 20))


# Identify matched rows
matched_rows <- inner_join(attention, gbd_long, by = "MAPPED_TRAIT_URI")
matched_rows 
(head(matched_rows, 20))



```


Verification and Checking for unique identifiers

```{r}
# Check the number of matched rows
num_matched_rows <- nrow(matched_rows)
(paste("Number of matched rows:", num_matched_rows))


# Check the number of unmatched rows in attention
num_unmatched_attention <- nrow(unmatched_attention)
(paste("Number of unmatched rows in attention:", num_unmatched_attention))

# Check the number of unmatched rows in gbd_long
num_unmatched_gbd <- nrow(unmatched_gbd)
(paste("Number of unmatched rows in gbd_long:", num_unmatched_gbd))


# Check for unique identifiers in the matched rows
unique_identifiers <- n_distinct(matched_rows$MAPPED_TRAIT_URI)
(paste("Number of unique identifiers in matched rows:", unique_identifiers))

# Check for unique identifiers in unmatched rows in gbd_long
gbd_long_unique_identifiers <- n_distinct(unmatched_gbd$MAPPED_TRAIT_URI)
(paste("Number of unique identifiers in unmatched rows in gbd_long:", gbd_long_unique_identifiers))

# Check for unique identifiers in the matched rows in attention
attention_unique_identifiers <- n_distinct(unmatched_attention$MAPPED_TRAIT_URI)
(paste("Number of unique identifiers in unmatched rows in attention:", attention_unique_identifiers))


```

save the files after removing the duplicates

```{r}
#  Removing the duplicates 
matched_rows <- inner_join(attention, gbd_long, by = "MAPPED_TRAIT_URI") %>%
  distinct(MAPPED_TRAIT_URI, .keep_all = TRUE)
unmatched_attention <- anti_join(attention, gbd_long, by = "MAPPED_TRAIT_URI") %>%
  distinct(MAPPED_TRAIT_URI, .keep_all = TRUE)
unmatched_gbd <- anti_join(gbd_long, attention, by = "MAPPED_TRAIT_URI") %>%
  distinct(MAPPED_TRAIT_URI, .keep_all = TRUE)



#check for any duplicates
duplicates <- unmatched_gbd[duplicated(unmatched_gbd$MAPPED_TRAIT_URI), ]
if (nrow(duplicates) > 0) {
  print("Duplicates found in unmatched_gbd:")
  print(duplicates)
} else {
  print("No duplicates found in unmatched_gbd.")
}

duplicates <- unmatched_attention[duplicated(unmatched_attention$MAPPED_TRAIT_URI), ]
if (nrow(duplicates) > 0) {
  print("Duplicates found in unmatched_attention:")
  print(duplicates)
} else {
  print("No duplicates found in unmatched_attention.")
}

duplicates <- matched_rows[duplicated(matched_rows$MAPPED_TRAIT_URI), ]
if (nrow(duplicates) > 0) {
  print("Duplicates found in matched rows:")
  print(duplicates)
} else {
  print("No duplicates found in matched rows.")
}


# save the files 

#write.csv(matched_rows, "matched_rows.csv", row.names = FALSE)
#write.csv(unmatched_attention, "unmatched_attention.csv", row.names = FALSE)
#write.csv(unmatched_gbd, "unmatched_gbd.csv", row.names = FALSE)

```


