---
title: "Example analysis of Rayan's project"
output: html_notebook
---


Read in the data

```{r}
# Load necessary libraries

library(dplyr)
library(data.table)
library(tidyr)
library(stringr)
library(DescTools)
library(ggplot2)
library(stats)
library(readr)
library(ontologyIndex)
library(readxl)
library(openxlsx)
library(boot)
library(LorenzRegression)


# a <- fread("~/Downloads/gwas_catalog_v1.0.2.1-studies_r2024-03-11.tsv")

a <- fread("/Users/rayanaloliet/Desktop/GWAS Catalog/gwas_catalog_v1.0.2.1-studies_r2024-06-07.csv")
 


```



Have a look at the data

```{r}
hist(a$`ASSOCIATION COUNT`, breaks=100)
```

Sometimes there is more than one EFO term (comma separated). Split them into separate lines

```{r}

head(a, 10)

a <- a %>%
    tidyr::separate_rows(MAPPED_TRAIT_URI, sep = ", ")

head(a, 10)

```

Get the number of EFO terms per publication

```{r}
pubmed_count <- a %>% group_by(PUBMEDID) %>% 
    summarise(
        n_efo = length(unique(MAPPED_TRAIT_URI))
    ) %>% arrange(desc(n_efo))

pubmed_count

 
```

Add the EFO count per PUBMEDID to the data frame

```{r}
a <- left_join(a, pubmed_count, by="PUBMEDID")
str(a)
```

Step 1. Develop the attention scores

- n = number of studies with that EFO term
- weighted_n = weighted by number of EFO terms in the study in total
- nhits = number of GWAS hits reported for that EFO term
- weighted_nhits = nhits weighted by number of EFO terms in the study in total

```{r}
attention <- a %>% group_by(
    MAPPED_TRAIT_URI
) %>% 
    summarise(
        n = n(), 
        weighted_n = sum(1/n_efo),
        nhits = sum(`ASSOCIATION COUNT`),
        weighted_nhits = sum(`ASSOCIATION COUNT` / n_efo)
    )

str(attention)

```

Removing extra whitespace, cleaning the Impact factor values and ensure the values are numeric 

```{r}
# Remove extra whitespace
a$`Impact factor` <- trimws(a$`Impact factor`)

# Replace non-standard characters
a$`Impact factor` <- gsub("·", ".", a$`Impact factor`)

# Convert to numeric, forcing non-numeric values to NA (for those with no impact factors)
a$`Impact factor` <- as.numeric(a$`Impact factor`)

# Replace NA values with 0
a$`Impact factor`[is.na(a$`Impact factor`)] <- 0

# Verify the changes
str(a)
unique(a$`Impact factor`)
summary(a$`Impact factor`)


```

 Including the impact factor values after preparing and cleaning the attention scores. Additionally, we have added the PUBMEDID to ensure that the studies are independent when we sum up the "n." We also added DISEASE/TRAIT for matching via str_detect function if we did not find a match via EFO terms. 
 
```{r}


# Including the impact factor and (`DISEASE/TRAIT`) column for matching and PUBMEDID to ensure the the studies are independent during the mapping process 

attention <- a %>%
  group_by(MAPPED_TRAIT_URI) %>%
  summarise(
    n = n(), 
    weighted_n = sum(1 / n_efo),
    nhits = sum(`ASSOCIATION COUNT`),
    weighted_nhits = sum(`ASSOCIATION COUNT` / n_efo),
    weighted_attention_score_impact_factor = sum((1 / n_efo) * `Impact factor`, na.rm = TRUE),
     DISEASE_TRAIT = paste(unique(a$`DISEASE/TRAIT`[a$MAPPED_TRAIT_URI == cur_group()$MAPPED_TRAIT_URI]), collapse = "; "),
    PUBMEDID = paste(unique(PUBMEDID), collapse = ", ")
  )


 

# Expand PUBMEDID into separate rows
attention <- attention %>%
  separate_rows(PUBMEDID, sep = ", ") 


str(attention)

```
 
Step 2: Merging the GWAS attention scores with the GBD disease burden results. The GBD terms were split into first and second parts because the terms in the second part have descendants. Therefore, those in the second part need to be handled differently, where we have to pull out the descendants from the tree via ontologyIndex library and then do the matching.


```{r}

#First_part_GBD<- fread("/Users/rayanaloliet/Desktop/GWAS Catalog/GBD.csv")
First_part_GBD<- read_xlsx("/Users/rayanaloliet/Desktop/GWAS Catalog/First_part_GBD.xlsx")

str(First_part_GBD)

  # unique matched rows  by GBD term
First_part_GBD<- First_part_GBD%>%
  distinct(`GBD term`, .keep_all = TRUE)

 
```


Specify columns that contain EFO terms

```{r}

efo_columns <- c("EFO 1", "EFO 2", "EFO 3", "EFO 4", "EFO 5", "EFO 6", 
                 "EFO 7", "EFO 8", "EFO 9", "EFO 10", "EFO 11", "EFO 12",
                 "EFO 13", "EFO 14", "EFO 15", "EFO 16", "EFO 17", "EFO 18",
                 "EFO 19", "EFO 20", "EFO 21", "EFO 22", "EFO 23", "EFO 24",
                 "EFO 25", "EFO 26", "EFO 27", "EFO 28", "EFO 29", "EFO 30")
```


Reshape the GBD data to gather all relevant columns into key-value pairs and drop rows with NA EFO terms

```{r}

 gbd_long <- First_part_GBD%>%
pivot_longer(cols = all_of(efo_columns), names_to = "EFO_number", values_to =  "MAPPED_TRAIT_URI") %>% drop_na(MAPPED_TRAIT_URI) 

gbd_long
```

Separate rows based on a separator (assuming comma-separated EFO terms) and Trim whitespace

```{r}
gbd_long <- gbd_long %>%
  separate_rows(MAPPED_TRAIT_URI, sep = ",") %>%
  mutate(MAPPED_TRAIT_URI = trimws(MAPPED_TRAIT_URI))  

gbd_long
```

Ensure consistent formatting, cleaning and standardize MAPPED_TRAIT_URI in gbd_long (GBD file)

*There are many functions here because each time I attempted to find a matching, I was unable to do so due to the differences in the two datasets. For example, in GWAS, the identifiers have underscores, while in GBD they have colons, and so on... Therefore, the functions below are intended to unify these differences. 

```{r}

# Ensure consistent formatting
gbd_long$MAPPED_TRAIT_URI <- tolower(trimws(gbd_long$MAPPED_TRAIT_URI))

# Ensure lowercase
gbd_long$MAPPED_TRAIT_URI <- tolower(gbd_long$MAPPED_TRAIT_URI) 

# Trim leading/trailing whitespace
gbd_long$MAPPED_TRAIT_URI <- trimws(gbd_long$MAPPED_TRAIT_URI) 

# Remove all spaces
gbd_long$MAPPED_TRAIT_URI <- gsub("\\s+", "", gbd_long$MAPPED_TRAIT_URI) 

# Remove all internal spaces explicitly
gbd_long$MAPPED_TRAIT_URI <- gsub(" ", "", gbd_long$MAPPED_TRAIT_URI) 

# Convert to lowercase and remove spaces 
gbd_long$MAPPED_TRAIT_URI <- tolower(gsub("\\s+", "", gbd_long$MAPPED_TRAIT_URI)) 

# Remove colons and spaces
gbd_long$MAPPED_TRAIT_URI <- gsub("[:\\s]", "", gbd_long$MAPPED_TRAIT_URI)  

# Remove quotation marks
gbd_long$MAPPED_TRAIT_URI <- gsub("\"", "", gbd_long$MAPPED_TRAIT_URI) 

# Replace _ with : 
gbd_long$MAPPED_TRAIT_URI <- gsub("_", ":", gbd_long$MAPPED_TRAIT_URI) 

# Remove all extra spaces
gbd_long$MAPPED_TRAIT_URI <- str_squish(gbd_long$MAPPED_TRAIT_URI) 

(head(gbd_long$MAPPED_TRAIT_URI, 20))

```

Ensure consistent formatting, cleaning and standardize MAPPED_TRAIT_URI in attention (GWAS file) and remove rows with NA in MAPPED_TRAIT_URI (EFO terms)

```{r}
# Function to clean and standardize MAPPED_TRAIT_URI
clean_mapped_trait_uri <- function(uri) {
  # Convert to lowercase
  uri <- tolower(uri)
  # Trim leading/trailing whitespace
  uri <- trimws(uri)
  # Remove all internal spaces explicitly
  uri <- gsub("\\s+", "", uri)
  # Remove quotation marks
  uri <- gsub("\"", "", uri)
  # Replace _ with :
  uri <- gsub("_", ":", uri)
  # Remove colons and spaces
  uri <- gsub("[:\\s]", "", uri)
  # Remove prefixes and ensure consistency
  uri <- sub(".*[:/]", "", uri)
  return(uri)
}

# Apply the function to MAPPED_TRAIT_URI column
attention$MAPPED_TRAIT_URI <- clean_mapped_trait_uri(attention$MAPPED_TRAIT_URI)


# Convert "na" strings to actual NA values
attention$MAPPED_TRAIT_URI[attention$MAPPED_TRAIT_URI == "na"] <- NA

# Remove rows with NA in MAPPED_TRAIT_URI
attention <- attention %>%
  filter(!is.na(MAPPED_TRAIT_URI))

(head(attention$MAPPED_TRAIT_URI, 20))

```

Finding a match between GWAS and GBD (First_part_GBD) via identifiers (EFO terms)- for those with no descendants. Then grouping those terms by GBD term and PUBMEDID (as we need to ensure the studies are independent when we sum the n ).  

*many-to-many relationship between `x` and `y` are expected due to grouping via GBD term and PUBMEDID.

```{r}

# Merge the data frames
merged_data <- merge(attention, gbd_long, by = "MAPPED_TRAIT_URI", all.x = TRUE)

# Identify matched rows
matched_rows <- inner_join(attention, gbd_long, by = "MAPPED_TRAIT_URI")

 
 # Group by GBD term and PUBMEDID and summarize within these groups 
 matched_rows_unique <- matched_rows %>%
    group_by(`GBD term`, PUBMEDID) %>%
    summarise(
      total_attention_score_per_pubmed = first(n, na.rm = TRUE),
      weighted_n = first(weighted_n),
      nhits = first(nhits),
      weighted_nhits = first(weighted_nhits),
      weighted_attention_score_impact_factor = first(weighted_attention_score_impact_factor),
      unique_pubmed_ids = n_distinct(PUBMEDID),
      MAPPED_TRAIT_URI = first(MAPPED_TRAIT_URI),
      DISEASE_TRAIT = first(DISEASE_TRAIT),
      .groups = "drop"
    )
 
      # Now summarize these attention scores across the GBD term and sum the total_attention_score_per_pubmed 
  unique_matched_gbd_terms_First_part_GBD <- matched_rows_unique %>%
    group_by(`GBD term`) %>%
    summarise(
      total_attention_score = sum(total_attention_score_per_pubmed, na.rm = TRUE),
      weighted_n = first(weighted_n),
      nhits = first(nhits),
      weighted_nhits = first(weighted_nhits),
      weighted_attention_score_impact_factor = first(weighted_attention_score_impact_factor),
      unique_pubmed_ids = n_distinct(PUBMEDID),
      MAPPED_TRAIT_URI = first(MAPPED_TRAIT_URI),
      DISEASE_TRAIT = first(DISEASE_TRAIT),
      .groups = "drop"
    )
 
   
# Identify unmatched rows from attention
unmatched_attention <- anti_join(attention, gbd_long, by = "MAPPED_TRAIT_URI")

# Identify unmatched rows from gbd_long
unmatched_gbd <- anti_join(gbd_long, attention, by = "MAPPED_TRAIT_URI")

# Ensure unique unmatched_gbd rows by GBD term
unmatched_gbd_unique <- unmatched_gbd %>%
  distinct(`GBD term`, .keep_all = TRUE)

# Extract unique GBD terms for matched and unmatched entries
matched_gbd_terms <- matched_rows_unique$`GBD term` %>% unique()
unmatched_gbd_terms <- unmatched_gbd_unique$`GBD term` %>% unique()

# Remove any GBD terms from unmatched_gbd that are also in matched_rows
unique_unmatched_gbd_terms <- setdiff(unmatched_gbd_terms, matched_gbd_terms)

# Ensure unique unmatched_gbd rows by GBD term
unique_unmatched_gbd_terms_First_part_GBD <- unmatched_gbd_unique %>%
  filter(`GBD term` %in% unique_unmatched_gbd_terms)

# Save the data frames to Excel files
#write.xlsx(unique_matched_gbd_terms_First_part_GBD, "unique_matched_gbd_terms_First_part_GBD.xlsx")
#write.xlsx(unique_unmatched_gbd_terms_First_part_GBD, "unique_unmatched_gbd_terms_First_part_GBD.xlsx")

```


Verification of the summing up of n in total_attention_score_per_pubmed "as long as they are independent publications" (all PUBMEDID are unique)

```{r}

# Select a specific GBD term for manual verification
selected_gbd_term <- "Acne vulgaris"  # Example term

# Filter the intermediate summarization for the selected GBD term
filtered_intermediate <- matched_rows_unique %>%
  filter(`GBD term` == selected_gbd_term)

# Print the filtered data for manual verification
print(filtered_intermediate)

# Sum the `n` values manually for verification
manual_sum_n <- sum(filtered_intermediate$total_attention_score_per_pubmed)
print(paste("Manual sum of n for", selected_gbd_term, ":", manual_sum_n))

```


Finding a match between GWAS and GBD (Second_part_GBD) via identifiers- for those with  descendants. We need to change the format of the identifiers again to get the descendants using the ontology from the tree then reverse it again to match it with GWAS (to ensure the consistency in the matching process). Also, as we did not part one, we will group them by GBD term and PUBMEDID. 


```{r}


# Define utility functions
format_identifier <- function(identifier) {
  formatted <- gsub(" ", "", identifier) # Remove any spaces
  formatted <- toupper(formatted) # Convert to uppercase
  formatted <- gsub("([A-Z]+)([0-9]+)", "\\1:\\2", formatted) # Add colon
  return(formatted)
}

revert_identifier <- function(identifier) {
  reverted <- gsub(":", "", identifier) # Remove colons
  return(tolower(reverted)) # Convert to lowercase
}

clean_mapped_trait_uri <- function(uri) {
  uri <- tolower(uri) # Convert to lowercase
  uri <- trimws(uri) # Trim leading/trailing whitespace
  uri <- gsub("\\s+", "", uri) # Remove all internal spaces explicitly
  uri <- gsub("\"", "", uri) # Remove quotation marks
  uri <- gsub("_", ":", uri) # Replace _ with :
  uri <- gsub("[:\\s]", "", uri) # Remove colons and spaces
  uri <- sub(".*[:/]", "", uri) # Remove prefixes and ensure consistency
  return(uri)
}

# Read data files
ontology <- get_OBO("/Users/rayanaloliet/Desktop/GWAS Catalog/efo-obo.txt")
Second_part_GBD <- read_excel("/Users/rayanaloliet/Desktop/GWAS Catalog/Second_part_GBD.xlsx")


# Extract and combine identifiers from Second_part_GBD
identifiers_list <- list(
  Second_part_GBD$`MAPPED_TRAIT_URI...2`,
  Second_part_GBD$`MAPPED_TRAIT_URI...3`,
  Second_part_GBD$`MAPPED_TRAIT_URI...4`,
  Second_part_GBD$`MAPPED_TRAIT_URI...5`,
  Second_part_GBD$`MAPPED_TRAIT_URI...6`,
  Second_part_GBD$`MAPPED_TRAIT_URI...7`,
  Second_part_GBD$`MAPPED_TRAIT_URI...8`,
  Second_part_GBD$`MAPPED_TRAIT_URI...9`,
  Second_part_GBD$`MAPPED_TRAIT_URI...10`,
  Second_part_GBD$`MAPPED_TRAIT_URI...11`,
  Second_part_GBD$`MAPPED_TRAIT_URI...12`,
  Second_part_GBD$`MAPPED_TRAIT_URI...13`
)
identifiers <- do.call(c, lapply(identifiers_list, na.omit))

# Format the identifiers
formatted_identifiers <- sapply(identifiers, format_identifier)

# Extract descendants for each identifier
descendants_list <- list()
for (i in 1:length(formatted_identifiers)) {
  id <- formatted_identifiers[i]
  if (id %in% ontology$id) {
    descendants <- get_descendants(ontology, id)
    descendants_list[[id]] <- descendants
  } else {
    descendants_list[[id]] <- NA
  }
}

# Determine the maximum number of descendants
max_descendants <- max(sapply(descendants_list, function(x) if (is.null(x) || all(is.na(x))) 0 else length(x)), na.rm = TRUE)

# Create a data frame for descendants
descendants_df <- data.frame(MAPPED_TRAIT_URI = character(), stringsAsFactors = FALSE)
for (i in 1:max_descendants) {
  descendants_df[paste0("Descendant_", i)] <- character()
}

# Populate the data frame with identifiers and their descendants
for (i in seq_along(formatted_identifiers)) {
  id <- formatted_identifiers[i]
  descendants <- descendants_list[[id]]
  if (is.null(descendants) || all(is.na(descendants))) {
    row <- c(MAPPED_TRAIT_URI = id, rep("", max_descendants))
  } else {
    row <- c(MAPPED_TRAIT_URI = id, descendants, rep("", max_descendants - length(descendants)))
  }
  
  row_df <- as.data.frame(t(row), stringsAsFactors = FALSE)
  colnames(row_df) <- colnames(descendants_df)
  descendants_df <- rbind(descendants_df, row_df)
}

# Convert columns to character type
descendants_df[] <- lapply(descendants_df, as.character)

# Add the GBD_TERM column
gbd_terms <- Second_part_GBD$`GBD term`
descendants_df$GBD_TERM <- gbd_terms[match(descendants_df$MAPPED_TRAIT_URI, formatted_identifiers)]

# Reorder columns to place GBD_TERM after MAPPED_TRAIT_URI
if ("GBD_TERM" %in% colnames(descendants_df)) {
  descendants_df <- descendants_df %>% select(MAPPED_TRAIT_URI, GBD_TERM, everything())
}

# Revert the identifiers to their original format
descendants_df <- descendants_df %>% mutate(across(c(MAPPED_TRAIT_URI, starts_with("Descendant_")), ~sapply(., revert_identifier)))

# Ensure identifiers are formatted correctly in attention
attention <- attention %>% mutate(MAPPED_TRAIT_URI = format_identifier(MAPPED_TRAIT_URI))

# Remove duplicates in both data frames
attention <- attention %>% distinct(MAPPED_TRAIT_URI, .keep_all = TRUE)
descendants_df <- descendants_df %>% distinct(MAPPED_TRAIT_URI, .keep_all = TRUE)

# Reshape descendants_df to long format
descendants_long <- descendants_df %>%
  pivot_longer(cols = starts_with("Descendant_"), names_to = "Descendant", values_to = "Descendant_URI") %>%
  filter(Descendant_URI != "") %>%
  select(MAPPED_TRAIT_URI, GBD_TERM, Descendant_URI)

# Clean and standardize Descendant_URI
descendants_long$Descendant_URI <- tolower(trimws(descendants_long$Descendant_URI))
descendants_long$Descendant_URI <- gsub("\\s+", "", descendants_long$Descendant_URI)
descendants_long$Descendant_URI <- gsub(" ", "", descendants_long$Descendant_URI)
descendants_long$Descendant_URI <- gsub("[:\\s]", "", descendants_long$Descendant_URI)
descendants_long$Descendant_URI <- gsub("\"", "", descendants_long$Descendant_URI)
descendants_long$Descendant_URI <- gsub("_", ":", descendants_long$Descendant_URI)
descendants_long$Descendant_URI <- str_squish(descendants_long$Descendant_URI)
descendants_long <- descendants_long %>% distinct(Descendant_URI, .keep_all = TRUE)

# Clean and standardize MAPPED_TRAIT_URI in attention
attention$MAPPED_TRAIT_URI <- clean_mapped_trait_uri(attention$MAPPED_TRAIT_URI)

# Identify matched entries
matched_entries <- attention %>%
  filter(MAPPED_TRAIT_URI %in% descendants_long$Descendant_URI) %>%
  left_join(descendants_long %>% select(Descendant_URI, GBD_TERM), by = c("MAPPED_TRAIT_URI" = "Descendant_URI"))


 # Group by GBD term and PUBMEDID and summarize within these groups and drop na 
 matched_entries_unique <- matched_entries %>% 
    drop_na() %>%
    group_by(GBD_TERM, PUBMEDID) %>%
    summarise(
      total_attention_score_per_pubmed = first(n, na.rm = TRUE),
      weighted_n = first(weighted_n),
      nhits = first(nhits),
      weighted_nhits = first(weighted_nhits),
      weighted_attention_score_impact_factor = first(weighted_attention_score_impact_factor),
      unique_pubmed_ids = n_distinct(PUBMEDID),
      MAPPED_TRAIT_URI = first(MAPPED_TRAIT_URI),
      DISEASE_TRAIT = first(DISEASE_TRAIT),
      .groups = "drop"
    )


     # Now summarize these attention scores across the GBD term and sum the total_attention_score_per_pubmed 
  unique_matched_gbd_terms_Second_part_GBD <- matched_entries_unique %>%
    group_by(GBD_TERM) %>%
    summarise(
      total_attention_score = sum(total_attention_score_per_pubmed, na.rm = TRUE),
      weighted_n = first(weighted_n),
      nhits = first(nhits),
      weighted_nhits = first(weighted_nhits),
      weighted_attention_score_impact_factor = first(weighted_attention_score_impact_factor),
      unique_pubmed_ids = n_distinct(PUBMEDID),
      MAPPED_TRAIT_URI = first(MAPPED_TRAIT_URI),
      DISEASE_TRAIT = first(DISEASE_TRAIT),
      .groups = "drop"
    )
  
# Identify unmatched GBD entries based on GBD_TERM in both datasets
# First, extract unique GBD terms from both datasets
matched_gbd_terms <- matched_entries$GBD_TERM %>% unique()
unmatched_gbd_terms <- Second_part_GBD$`GBD term` %>% unique()

# Identify unmatched GBD terms
unique_unmatched_gbd_terms <- setdiff(unmatched_gbd_terms, matched_gbd_terms)

# Filter the Second_part_GBD dataframe to get rows with these unique unmatched GBD terms
unique_unmatched_gbd_terms_Second_part_GBD <- Second_part_GBD %>%
  filter(`GBD term` %in% unique_unmatched_gbd_terms)

# Save the data frames to Excel files
#write.xlsx(unique_unmatched_gbd_terms_Second_part_GBD, unique_unmatched_gbd_terms_Second_part_GBD.xlsx")
#write.xlsx(unique_matched_gbd_terms_Second_part_GBD, "unique_matched_gbd_terms_Second_part_GBD.xlsx")

```

Combined matched term and unmatched terms in one data set: 
unique_matched_gbd_terms_First_part_GBD & unique_matched_gbd_terms_Second_part_GBD (140)
unique_unmatched_gbd_terms_First_part_GBD & unique_unmatched_gbd_terms_Second_part_GBD (237)

```{r}

# List of columns to keep for unmatched terms
columns_unmatched <- c("GBD term", "EFO_number", "MAPPED_TRAIT_URI")

# Select only the necessary columns from the first unmatched data frame
unique_unmatched_gbd_terms_First_part_GBD <- unique_unmatched_gbd_terms_First_part_GBD[, columns_unmatched]

# Add missing columns to the second unmatched data frame
if (!"EFO_number" %in% names(unique_unmatched_gbd_terms_Second_part_GBD)) {
  unique_unmatched_gbd_terms_Second_part_GBD$EFO_number <- NA
}

# Rename the appropriate column in the second unmatched data frame
names(unique_unmatched_gbd_terms_Second_part_GBD)[which(names(unique_unmatched_gbd_terms_Second_part_GBD) == "MAPPED_TRAIT_URI...2")] <- "MAPPED_TRAIT_URI"

# Select and order the columns to match the first unmatched data frame
unique_unmatched_gbd_terms_Second_part_GBD <- unique_unmatched_gbd_terms_Second_part_GBD[, columns_unmatched]

# Combine the unmatched terms
combined_unmatched_terms <- rbind(unique_unmatched_gbd_terms_First_part_GBD, unique_unmatched_gbd_terms_Second_part_GBD)

# Display the results
print("Combined Unmatched Terms:")
print(combined_unmatched_terms)


# List of columns to keep for matched terms
columns_matched <- c("MAPPED_TRAIT_URI", "total_attention_score", "weighted_n", "nhits", "weighted_nhits", "weighted_attention_score_impact_factor", "unique_pubmed_ids", "GBD term", "DISEASE_TRAIT")

# Select only the necessary columns from the first matched data frame
unique_matched_gbd_terms_First_part_GBD <- unique_matched_gbd_terms_First_part_GBD[, columns_matched]

# Rename the column in the second matched data frame
names(unique_matched_gbd_terms_Second_part_GBD)[which(names(unique_matched_gbd_terms_Second_part_GBD) == "GBD_TERM")] <- "GBD term"

# Ensure all columns are present in the second matched data frame, filling with NA where necessary
for (col in columns_matched) {
  if (!col %in% names(unique_matched_gbd_terms_Second_part_GBD)) {
    unique_matched_gbd_terms_Second_part_GBD[[col]] <- NA
  }
}

# Select and order the columns to match the first matched data frame
unique_matched_gbd_terms_Second_part_GBD <- unique_matched_gbd_terms_Second_part_GBD[, columns_matched]

# Combine the matched terms
combined_matched_terms <- rbind(unique_matched_gbd_terms_First_part_GBD, unique_matched_gbd_terms_Second_part_GBD)

# Display the results
print("Combined Matched Terms:")
print(combined_matched_terms)

# Save the data frames to Excel files
#write.xlsx(combined_matched_terms, combined_matched_terms.xlsx")
#write.xlsx(combined_unmatched_terms, "combined_unmatched_terms.xlsx")


```

Here we could follow a partial matching approach to handle unmatched GBD terms using "str_detect" function to detect for matching but with different identifiers or GBD present within a string of traits in GWAS.

*This approach is helpful because sometimes we may miss EFO terms in the mapping process during manual mapping i.e there are many EFO terms for a particular GBD, this will be capture using str_detect" function.

To ensure consistent formatting and standardize of DISEASE/TRAIT in unmatched GBD terms and unmatched attention (GWAS) ie., both lower case for comparison, we will create new columns in both data sets 


```{r}
# Create a new column for disease names to facilitate partial matching in attention
attention <- attention %>%
  mutate(MAPPED_TRAIT_NAME = tolower(trimws(gsub("[^a-zA-Z0-9 ]", "", DISEASE_TRAIT))))

# Create a new column for disease names to facilitate partial matching in combined_unmatched_terms
combined_unmatched_terms <- combined_unmatched_terms%>%
  mutate(GBD_TERM_NAME = tolower(trimws(gsub("[^a-zA-Z0-9 ]", "", combined_unmatched_terms$`GBD term`))))


# Perform partial matching between MAPPED_TRAIT_NAME in attention and GBD_TERM_NAME in combined_unmatched_terms with the rest of the information 

partial_matches <- combined_unmatched_terms %>%
  cross_join(attention) %>%
  filter(str_detect(MAPPED_TRAIT_NAME, GBD_TERM_NAME)) %>%
  select(
    GBD_TERM_NAME, 
    MAPPED_TRAIT_NAME, 
    GBD_TRAIT_URI = MAPPED_TRAIT_URI.x, 
    MAPPED_TRAIT_URI = MAPPED_TRAIT_URI.y,
    n, 
    weighted_n, 
    nhits, 
    weighted_nhits, 
    weighted_attention_score_impact_factor,
    PUBMEDID,
    `GBD term`
  ) %>%
  distinct()



# Group by GBD term and PUBMEDID and summarize within these groups 
 partial_matches_filtered <- partial_matches %>%
    group_by(`GBD term`, PUBMEDID) %>%
    summarise(
      total_attention_score_per_pubmed = first(n, na.rm = TRUE),
      weighted_n = first(weighted_n),
      nhits = first(nhits),
      weighted_nhits = first(weighted_nhits),
      weighted_attention_score_impact_factor = first(weighted_attention_score_impact_factor),
      unique_pubmed_ids = n_distinct(PUBMEDID),
      MAPPED_TRAIT_URI = first(MAPPED_TRAIT_URI),
      GBD_TRAIT_URI = first(GBD_TRAIT_URI),
      MAPPED_TRAIT_NAME = first(MAPPED_TRAIT_NAME),
      .groups = "drop"
    )
 
     # Now summarize these attention scores across the GBD term and sum the total_attention_score_per_pubmed 
  partial_matches_unique <- partial_matches_filtered %>%
    group_by(`GBD term`) %>%
    summarise(
      total_attention_score = sum(total_attention_score_per_pubmed, na.rm = TRUE),
      weighted_n = first(weighted_n),
      nhits = first(nhits),
      weighted_nhits = first(weighted_nhits),
      weighted_attention_score_impact_factor = first(weighted_attention_score_impact_factor),
      unique_pubmed_ids = n_distinct(PUBMEDID),
      MAPPED_TRAIT_URI = first(MAPPED_TRAIT_URI),
      GBD_TRAIT_URI = first(GBD_TRAIT_URI),
      MAPPED_TRAIT_NAME = first(MAPPED_TRAIT_NAME),
      .groups = "drop"
    )



```


Identify the GBD terms that are not mapped via identifiers or trait with GWAS to combine them
in one data set. Now we will have final_matched_gbd and final_unmatched_gbd. 

```{r}


# Identify the rows that were matched partially
matched_partial_gbd <- partial_matches_unique %>% 
  select(GBD_TRAIT_URI, `GBD term`) %>% 
  distinct() 

# Get final unmatched rows
final_unmatched_gbd <- anti_join(combined_unmatched_terms, matched_partial_gbd, by = "GBD term")


# Check for duplicates in final_unmatched_gbd
duplicates_final_unmatched_gbd <- final_unmatched_gbd[duplicated(final_unmatched_gbd$`GBD term`), ]
if (nrow(duplicates_final_unmatched_gbd) > 0) {
  print("Duplicates found in final_unmatched_gbd:")
  print(duplicates_final_unmatched_gbd)
} else {
  print("No duplicates found in final_unmatched_gbd.")
}


# final matched_gbd (combine those identified via Identifier and Partially)
final_matched_gbd <- bind_rows(
  combined_matched_terms %>% mutate(Match_Type = "Identifier"),
  partial_matches_unique %>% mutate(Match_Type = "Partial")
)

# Check for duplicates in final matched_gbd
duplicates_final_matched_gbd <- final_matched_gbd[duplicated(final_matched_gbd$`GBD term`), ]
if (nrow(duplicates_final_matched_gbd) > 0) {
  print("Duplicates found in final_matched_gbd:")
  print(duplicates_final_matched_gbd)
} else {
  print("No duplicates found in final_matched_gbd")
}


# Save the results to a CSV file
#write.csv(final_unmatched_gbd, "final_unmatched_gbd.csv", row.names = FALSE)
#write.csv(final_matched_gbd, "final_matched_gbd.csv", row.names = FALSE)


```


If there are no matches for a GBD trait then the attention score = 0. also, we ill combine matced and unmatched GBD terms in one data set (combined dataset). It has 377 GBD terms, as we started. 

```{r}


#  Add 'total_attention_score' column with value 0 to final_unmatched_gbd
final_unmatched_gbd <- final_unmatched_gbd %>%
  mutate(total_attention_score = 0)

#  Ensure both datasets have the same columns
# Here, we ensure that final_unmatched_gbd has all necessary columns
required_columns <- names(final_matched_gbd)

# Add any missing columns to final_unmatched_gbd and set their values to NA
for (col in required_columns) {
  if (!(col %in% names(final_unmatched_gbd))) {
    final_unmatched_gbd[[col]] <- NA
  }
}

# Combine the datasets
combined_dataset <- bind_rows(final_matched_gbd, final_unmatched_gbd)

combined_dataset <- combined_dataset %>%
  distinct(`GBD term`, .keep_all = TRUE)


# Save the combined dataset to a CSV file
#write.csv(combined_dataset, "combined_dataset_EFOU.csv", row.names = FALSE)

```


load the GBD DALY data, verify any missing values, and inspect the structure of the dataset
 
```{r}

# Load the GBD DALY data
gbd_daly_data <- read_csv('/Users/rayanaloliet/Desktop/GWAS Catalog/IHME-GBD_2021_DATA-120ebfcd-1.csv')

# Filter out rows with NA values in the key columns
gbd_daly_data <- gbd_daly_data %>%
  filter(!is.na(`cause_name`), !is.na(val))

# Verify that there are no NA values left
sum(is.na(gbd_daly_data$`cause_name`))
sum(is.na(gbd_daly_data$val))
sum(is.na(gbd_daly_data$measure_name))

 
# Inspect the structure of the GBD DALY data
str(gbd_daly_data)




```


Assessing the relationship between GWAS attention and global need, using an inequality measure.This will be done via filtering the data by Number of DALY and then merging by GBD term.

```{r}
#2019 
#gbd_daly_data <- gbd_daly_data %>%
 # filter(metric_name == "Number" & measure_name == "DALYs (Disability-Adjusted Life Years)") #%>%
 # rename(`GBD term`= cause_name, DALY = val)
#2021
gbd_daly_data <- gbd_daly_data %>%
  filter(metric_name == "Number") %>%
  rename(`GBD term`= cause_name, DALY = val)


# Merge the DALY values into the combined dataset
combined_dataset <- combined_dataset %>%
  left_join(gbd_daly_data, by = c("GBD term" = "GBD term"))


# Verify the merge
head(combined_dataset)

# Remove specified columns
combined_dataset <- combined_dataset %>%
  select(-c(MAPPED_TRAIT_URI, MAPPED_TRAIT_NAME, EFO_number, GBD_TERM_NAME))

# remove na
combined_dataset <- combined_dataset[!is.na(combined_dataset$location_name), ]

# Verify the data
summary(combined_dataset$total_attention_score)
summary(combined_dataset$DALY)

#write.csv(combined_dataset, "combined_dataset_EFOM.csv", row.names = FALSE)

```






Try to identify any difference or duplicates between combined dataset and the data set that we started with in the manual mapping.  
 

```{r}
unmatched_gbd_terms <- anti_join(gbd_daly_data, combined_dataset, by = c("GBD term" = "GBD term"))

unmatched_gbd_terms <- unmatched_gbd_terms %>%
  distinct(`GBD term`, .keep_all = TRUE)

# Get the unique GBD terms in each dataset
combined_gbd_terms <- unique(combined_dataset$`GBD term`)
gbd_gbd_terms <- unique(gbd_daly_data$`GBD term`)

# Identify terms in combined_dataset not in gbd_daly_data
terms_only_in_combined <- setdiff(combined_gbd_terms, gbd_gbd_terms)

# Identify terms in gbd_daly_data not in combined_dataset
terms_only_in_gbd_daly <- setdiff(gbd_gbd_terms, combined_gbd_terms)

# Get the lengths of the unique terms
len_combined <- length(terms_only_in_combined)
len_gbd_daly <- length(terms_only_in_gbd_daly)

# Determine the maximum length to ensure both columns in the data frame have the same number of rows
max_length <- max(len_combined, len_gbd_daly)

# Combine the differences into a data frame for better visualization
diff_2 <- data.frame(
  Terms_Only_in_Combined_Dataset = c(terms_only_in_combined, rep(NA, max_length - len_combined)),
  Terms_Only_in_GBD_Daly_Dataset = c(terms_only_in_gbd_daly, rep(NA, max_length - len_gbd_daly))
)

# the differences
(diff_2)

#write.csv(diff_2, "diff_2.csv", row.names = FALSE)
#write.csv(combined_dataset, "combined_dataset_EFO.csv", row.names = FALSE)


```


exclude GBD teams that are classified as Injuries and total 


```{r}
# List of injuries to exclude
terms_to_exclude <- c("Injuries", "Transport injuries", "Road injuries", 
                         "Pedestrian road injuries", "Cyclist road injuries", 
                         "Motorcyclist road injuries", "Motor vehicle road injuries", 
                         "Other road injuries", "Other transport injuries", 
                         "Unintentional injuries", "Falls", "Drowning", 
                         "Fire, heat, and hot substances", "Poisonings", 
                         "Poisoning by carbon monoxide", "Poisoning by other means", 
                         "Exposure to mechanical forces", "Unintentional firearm injuries", 
                         "Other exposure to mechanical forces", "Adverse effects of medical treatment", 
                         "Animal contact", "Venomous animal contact", 
                         "Non-venomous animal contact", "Foreign body", 
                         "Pulmonary aspiration and foreign body in airway", 
                         "Foreign body in eyes", "Foreign body in other body part", 
                         "Environmental heat and cold exposure", "Exposure to forces of nature", 
                         "Other unintentional injuries", "Self-harm and interpersonal violence", 
                         "Self-harm", "Self-harm by firearm", "Self-harm by other specified means", 
                         "Interpersonal violence", "Physical violence by firearm", 
                         "Physical violence by sharp object", "Sexual violence", 
                         "Physical violence by other means", "Conflict and terrorism", 
                         "Police conflict and executions", "All causes")

 # Filter the GBD DALY data for "Number" metric and rename columns
filtered_dataset <- combined_dataset %>%
  filter(!(`GBD term` %in% terms_to_exclude))  # Exclude specified terms

```



split the dataset into communicable and noncommunicable for sensitivity analysis
```{r}
broader_categories_dataset<- read_xlsx("/Users/rayanaloliet/Desktop/untitled folder 7/GBD_2021_CAUSE_HIERARCHY_Y2024M05D16.xlsx")
 
head(broader_categories_dataset)
```

```{r}
filtered_dataset <- filtered_dataset %>%
  rename(`Cause Name` = `GBD term`)

```

```{r}
merged_dataset <- merge(filtered_dataset, broader_categories_dataset, by = "Cause Name", all.x = TRUE)
```

based on the GBD, here we should have 95 * 27 locations for the Communicable, maternal, neonatal, and nutritional diseases
```{r}
merged_dataset_EFO_CMNN <- merged_dataset %>%
  filter(substr(`Cause Outline`, 1, 1) %in% c("A"))

write.csv(merged_dataset_EFO_CMNN, "merged_dataset_EFO_CMNN.csv", row.names = FALSE)
```

based on the GBD, here we should have 234 * 27 for Non-communicable diseases diseases
```{r}
merged_dataset_EFO_NC <- merged_dataset %>%
  filter(substr(`Cause Outline`, 1, 1) %in% c("B"))

write.csv(merged_dataset_EFO_NC, "merged_dataset_EFO_NC.csv", row.names = FALSE)
```


```{r}
merged_dataset_EFO_exclude_Injuries <- filtered_dataset

write.csv(merged_dataset_EFO_exclude_Injuries, "merged_dataset_EFO_exclude_Injuries.csv", row.names = FALSE)
```

